---
title: "Pairwise colocalization with colocPipeline"
author: "pecotmr authors"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pairwise colocalization with colocPipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    message = FALSE,
    warning = FALSE,
    fig.width = 7,
    fig.height = 4.5
)
```

## Overview

`colocPipeline()` runs pairwise QTL ↔ GWAS colocalization using
`coloc::coloc.bf_bf` on the SuSiE log Bayes-factor (LBF) matrices
carried by each fine-mapping entry. For multi-trait colocalization
across many QTL outcomes in one model, use
[`colocboostPipeline`](colocboost-pipeline.html) instead.

Inputs:

-  **QTL input** — a `QtlFineMappingResult` produced by
    [`fineMappingPipeline`](fine-mapping.html).
- **GWAS input** — either a `GwasSumStats` (fine-mapping run internally) or a
`GwasFineMappingResult` (used directly).

The output is a `ColocResult`: one element per tested
`(QTL credible set, GWAS credible set, block)` pair, holding that pair's
aligned variants. Pair-level posteriors (`PP.H0.abf` … `PP.H4.abf`) and the
per-variant `SNP.PP.H4` are both kept, and four accessors project the object
to whichever granularity an analysis needs — see
[Reading the result](#reading-the-result).

```{r load-packages}
library(pecotmr)
```

## Bundled inputs

```{r bundled}
data(qtlDatasetExample, gwasSumStatsS4Example)

# Build a QtlFineMappingResult from the bundled QtlDataset.
qtlFmr <- fineMappingPipeline(
    qtlDatasetExample,
    methods = "susie",
    cisWindow = 1e6
)
```

## QTL FMR + GWAS sumstats (inline GWAS fine-mapping)

The recommend approach is to provide `colocPipeline()` a `QtlFineMappingResult`
and a `GwasSumStats`. The pipeline runs `fineMappingPipeline()` internally on
the GWAS sumstats (with `finemappingMethods`) over the exact window used in the
`QtlFineMappingResult`, then runs `coloc.bf_bf` per (QTL, GWAS) pair.

```{r coloc-quickstart}
res <- colocPipeline(
    qtlFineMappingResult = qtlFmr,
    gwasInput = gwasSumStatsS4Example,
    finemappingMethods = "susie"
)
res
```

Use `returnGwasFineMapping = TRUE` to get the inline GWAS FMR back as an
attribute on the result:

```{r coloc-return-fmr, eval=FALSE}
res2 <- colocPipeline(
    qtlFineMappingResult = qtlFmr,
    gwasInput = gwasSumStatsS4Example,
    returnGwasFineMapping = TRUE
)
attr(res2, "gwasFineMapping")
```

If the QTL `FineMappingResult` has an LD sketch because it was run on summary
statistics it must match the LD sketch for the `GWASSumStats`.

## QTL FMR + pre-computed GWAS FMR

If you already have a `GwasFineMappingResult` (typically computed over LD
blocks), you can use that fine-mapping result directly:

```{r coloc-precomputed-fmr, eval=FALSE}
gwasFmr <- fineMappingPipeline(gwasSumStatsS4Example, methods = "susie")
colocPipeline(qtlFineMappingResult = qtlFmr, gwasInput = gwasFmr)
```

The disadvantage of this approach is that LD blocks will not align with cis-QTL
windows.  Ultimately this means that the variable selection in the QTL and GWAS
fine-mapping will have been done over different variant sets, which may reduce
the sensitivity for colocalizations, particuarly in areas with complex LD.

## Filtering effects fed into coloc

`coloc::coloc.bf_bf` accepts one LBF matrix for the QTL and one for hte GWAS.
By default, every effect with non-trivial prior variance (`V > priorTol`) gets
included. Two opt-in filters can adjust this:

| Filter | Description | Default |
|---|---|---|
| `filterLbfCs` | TRUE keeps only effects that produced a credible set (`trimmedFit$sets$cs_index`). | FALSE |
| `filterLbfCsSecondary` | Numeric coverage in (0,1). Runs a CS-concentration filter at that coverage. | NULL |
| `filterLbfCsConcentration` | Numeric in (0,1). Concentration factor: a CS at coverage X is kept only if it spans fewer than `nVariants * X * concentration` variants. | 0.5 |
| `priorTol` | Drop effects whose estimated prior variance is at or below this threshold. | 1e-9 |

```{r coloc-filter-example, eval=FALSE}
colocPipeline(
    qtlFineMappingResult = qtlFmr,
    gwasInput = gwasSumStatsS4Example,
    filterLbfCsSecondary = 0.5, # 50% CS concentration filter
    filterLbfCsConcentration = 0.5
) # span < 25% of locus variants
```

## Enrichment-aware coloc (enloc mode)

Pass a per-`(gwasStudy, qtlContext)` enrichment table (the output of
`qtlEnrichmentPipeline()`) to scale the shared-signal prior:

\[ p_{12}^{\text{used}} = \min\bigl(p_{12}\,(1 + \text{enrichment}),\,
p_{12,\max}\bigr) \]

Pairs without a matching enrichment row fall back to the baseline
`p12` with a warning. Two extra columns are added to the output:
`enrichment` (the lookup) and `p12Used` (what was actually passed to
`coloc.bf_bf`).

```{r coloc-enloc, eval=FALSE}
enrichmentTable <- data.frame(
    gwasStudy = "trait1",
    qtlContext = "brain",
    enrichment = 2.5,
    stringsAsFactors = FALSE
)
colocPipeline(
    qtlFineMappingResult = qtlFmr,
    gwasInput = gwasSumStatsS4Example,
    enrichment = enrichmentTable,
    p12Max = 1e-3
)
```

## PIP renormalization on variant intersection

When the QTL and GWAS fine-mapping results were fit on different
variant sets (common when the user declines RAISS imputation, or when
the GWAS FMR carries variants the QTL FMR doesn't), `adjustPips = TRUE`
(default) renormalizes the PIPs for each objectl to the cross-FMR variant
intersection before any per-pair inference. Pass `FALSE` to use the
FMRs as supplied.

```{r coloc-adjustpips, eval=FALSE}
colocPipeline(
    qtlFineMappingResult = qtlFmr,
    gwasInput = gwasSumStatsS4Example,
    adjustPips = TRUE
) # default
```

## Common parameters

```{r coloc-params, eval=FALSE}
colocPipeline(
    qtlFineMappingResult = qtlFmr,
    gwasInput = gwasSumStatsS4Example,
    filterLbfCs = FALSE,
    filterLbfCsSecondary = NULL, # e.g. 0.5
    filterLbfCsConcentration = 0.5,
    priorTol = 1e-9,
    p1 = 1e-4,
    p2 = 1e-4,
    p12 = 5e-6,
    finemappingMethods = "susie", # inline GWAS FM when needed
    returnGwasFineMapping = FALSE,
    enrichment = NULL, # data.frame for enloc mode
    p12Max = 1e-3,
    adjustPips = TRUE
)
```

## Reading the result

Nothing is filtered when the result is written, so every narrowing below is a
view decision the caller makes — and can revisit — rather than something baked
in at run time.

### Pairs

The native `coloc.bf_bf` granularity: one row per tested pair. Every testable
pair is reported verbatim, including both halves of a credible set that
straddles a block boundary.

```{r read-pairs}
pairs <- getColocPairs(res)
head(pairs[order(pairs$PP.H4.abf, decreasing = TRUE), ])
```

`qtlCs` / `gwasCs` are the fine-mapping effect indices, which are stable across
blocks; coloc's own `idx1` / `idx2` number the rows of a single call and are
not comparable between them. `qtlRetainedMass` / `gwasRetainedMass` report how
much of each effect's posterior survived reconciliation to the shared variant
set — a low value means the effect rested largely on variants the other side
does not carry.

### Variants

The per-variant layer: `colocPp` is `PP.H4.abf * SNP.PP.H4`, the posterior that
this particular variant is the shared causal one.

```{r read-variants}
variants <- getColocVariants(res)
head(variants[order(variants$colocPp, decreasing = TRUE), ])
```

### Credible sets

The coloc credible set is the smallest set of variants whose cumulative
`SNP.PP.H4` reaches `coverage`. It is a *third* variant set — not guaranteed to
be a subset of either input credible set — so its purity is recomputed from LD
rather than inherited. Filtering happens before any LD extraction, so a
stricter threshold costs strictly less.

```{r read-cs}
head(getColocCredibleSets(res, coverage = 0.95, minPp4 = 0.1))
```

### Genes

Pooled to one row per gene. Within a QTL credible set the per-block and
per-GWAS-credible-set results are mutually exclusive — the shared causal
variant is in one block or the other — so they are summed; distinct QTL
credible sets are independent signals and combine as `1 - prod(1 - p)`.

```{r read-genes}
getColocGenes(res)
```

The warning above is the diagnostic that rule produces: a within-credible-set
sum above 1 means several GWAS credible sets are competing for one QTL signal.
The value is clipped to 1, and the pair view shows which pairs were involved.

If you need the flat table the pipeline used to return, `as.data.frame(res)`
gives the pair view.

# Session info

```{r session-info}
sessionInfo()
```
