---
title: "Pairwise colocalization with colocPipeline"
author: "pecotmr authors"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pairwise colocalization with colocPipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    message = FALSE,
    warning = FALSE,
    fig.width = 7,
    fig.height = 4.5
)
```

## Overview

`colocPipeline()` runs pairwise QTL ↔ GWAS colocalization using
`coloc::coloc.bf_bf` on the SuSiE log Bayes-factor (LBF) matrices
carried by each fine-mapping entry. For multi-trait colocalization
across many QTL outcomes in one model, use
[`colocboostPipeline`](colocboost-pipeline.html) instead.

Inputs:

-  **QTL input** — a `QtlFineMappingResult` produced by
    [`fineMappingPipeline`](fine-mapping.html).
- **GWAS input** — either a `GwasSumStats` (fine-mapping run internally) or a
`GwasFineMappingResult` (used directly).

The output is a long-format data frame with one row per
`(QTL tuple, GWAS tuple, credible-set pair)` combination and the
standard coloc posterior columns `PP.H0.abf` … `PP.H4.abf`.

```{r load-packages}
library(pecotmr)
```

## Bundled inputs

```{r bundled}
data(qtlDatasetExample, gwasSumStatsS4Example)

# Build a QtlFineMappingResult from the bundled QtlDataset.
qtlFmr <- fineMappingPipeline(
    qtlDatasetExample,
    methods = "susie",
    cisWindow = 1e6
)
```

## QTL FMR + GWAS sumstats (inline GWAS fine-mapping)

The recommend approach is to provide `colocPipeline()` a `QtlFineMappingResult`
and a `GwasSumStats`. The pipeline runs `fineMappingPipeline()` internally on
the GWAS sumstats (with `finemappingMethods`) over the exact window used in the
`QtlFineMappingResult`, then runs `coloc.bf_bf` per (QTL, GWAS) pair.

```{r coloc-quickstart}
res <- colocPipeline(
    qtlFineMappingResult = qtlFmr,
    gwasInput = gwasSumStatsS4Example,
    finemappingMethods = "susie"
)
head(res)
```

Use `returnGwasFineMapping = TRUE` to get the inline GWAS FMR back as an
attribute on the result:

```{r coloc-return-fmr, eval=FALSE}
res2 <- colocPipeline(
    qtlFineMappingResult = qtlFmr,
    gwasInput = gwasSumStatsS4Example,
    returnGwasFineMapping = TRUE
)
attr(res2, "gwasFineMapping")
```

If the QTL `FineMappingResult` has an LD sketch because it was run on summary
statistics it must match the LD sketch for the `GWASSumStats`.

## QTL FMR + pre-computed GWAS FMR

If you already have a `GwasFineMappingResult` (typically computed over LD
blocks), you can use that fine-mapping result directly:

```{r coloc-precomputed-fmr, eval=FALSE}
gwasFmr <- fineMappingPipeline(gwasSumStatsS4Example, methods = "susie")
colocPipeline(qtlFineMappingResult = qtlFmr, gwasInput = gwasFmr)
```

The disadvantage of this approach is that LD blocks will not align with cis-QTL
windows.  Ultimately this means that the variable selection in the QTL and GWAS
fine-mapping will have been done over different variant sets, which may reduce
the sensitivity for colocalizations, particuarly in areas with complex LD.

## Filtering effects fed into coloc

`coloc::coloc.bf_bf` accepts one LBF matrix for the QTL and one for hte GWAS.
By default, every effect with non-trivial prior variance (`V > priorTol`) gets
included. Two opt-in filters can adjust this:

| Filter | Description | Default |
|---|---|---|
| `filterLbfCs` | TRUE keeps only effects that produced a credible set (`trimmedFit$sets$cs_index`). | FALSE |
| `filterLbfCsSecondary` | Numeric coverage in (0,1). Runs a CS-concentration filter at that coverage. | NULL |
| `filterLbfCsConcentration` | Numeric in (0,1). Concentration factor: a CS at coverage X is kept only if it spans fewer than `nVariants * X * concentration` variants. | 0.5 |
| `priorTol` | Drop effects whose estimated prior variance is at or below this threshold. | 1e-9 |

```{r coloc-filter-example, eval=FALSE}
colocPipeline(
    qtlFineMappingResult = qtlFmr,
    gwasInput = gwasSumStatsS4Example,
    filterLbfCsSecondary = 0.5, # 50% CS concentration filter
    filterLbfCsConcentration = 0.5
) # span < 25% of locus variants
```

## Enrichment-aware coloc (enloc mode)

Pass a per-`(gwasStudy, qtlContext)` enrichment table (the output of
`qtlEnrichmentPipeline()`) to scale the shared-signal prior:

\[ p_{12}^{\text{used}} = \min\bigl(p_{12}\,(1 + \text{enrichment}),\,
p_{12,\max}\bigr) \]

Pairs without a matching enrichment row fall back to the baseline
`p12` with a warning. Two extra columns are added to the output:
`enrichment` (the lookup) and `p12Used` (what was actually passed to
`coloc.bf_bf`).

```{r coloc-enloc, eval=FALSE}
enrichmentTable <- data.frame(
    gwasStudy = "trait1",
    qtlContext = "brain",
    enrichment = 2.5,
    stringsAsFactors = FALSE
)
colocPipeline(
    qtlFineMappingResult = qtlFmr,
    gwasInput = gwasSumStatsS4Example,
    enrichment = enrichmentTable,
    p12Max = 1e-3
)
```

## PIP renormalization on variant intersection

When the QTL and GWAS fine-mapping results were fit on different
variant sets (common when the user declines RAISS imputation, or when
the GWAS FMR carries variants the QTL FMR doesn't), `adjustPips = TRUE`
(default) renormalizes the PIPs for each objectl to the cross-FMR variant
intersection before any per-pair inference. Pass `FALSE` to use the
FMRs as supplied.

```{r coloc-adjustpips, eval=FALSE}
colocPipeline(
    qtlFineMappingResult = qtlFmr,
    gwasInput = gwasSumStatsS4Example,
    adjustPips = TRUE
) # default
```

## Common parameters

```{r coloc-params, eval=FALSE}
colocPipeline(
    qtlFineMappingResult = qtlFmr,
    gwasInput = gwasSumStatsS4Example,
    filterLbfCs = FALSE,
    filterLbfCsSecondary = NULL, # e.g. 0.5
    filterLbfCsConcentration = 0.5,
    priorTol = 1e-9,
    p1 = 1e-4,
    p2 = 1e-4,
    p12 = 5e-6,
    finemappingMethods = "susie", # inline GWAS FM when needed
    returnGwasFineMapping = FALSE,
    enrichment = NULL, # data.frame for enloc mode
    p12Max = 1e-3,
    adjustPips = TRUE
)
```

## Reading the result

```{r read-result}
str(res, max.level = 1L)
head(res[order(res$PP.H4.abf, decreasing = TRUE), ])
```

Each row is one CS-pair combination; in particular the `idx1` /
`idx2` columns identify the QTL / GWAS effect indices within the
respective LBF matrices, and `PP.H4.abf` is the standard coloc
shared-signal posterior.
