---
title: "Building QtlSumStats and GwasSumStats objects"
author: "pecotmr authors"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Building QtlSumStats and GwasSumStats objects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5)
hasVA <- requireNamespace("VariantAnnotation", quietly = TRUE)
```

## Overview

`GwasSumStats` and `QtlSumStats` are `DFrame`-subclass collections of
summary statistics. Each **row** carries one `entry` — a `GRanges` of
per-variant statistics — annotated by an identity tuple:

| Class | Row key | One row is |
|---|---|---|
| `GwasSumStats` | `study` | one GWAS study over a region |
| `QtlSumStats` | `(study, context, trait)` | one molecular trait in one context |

Every `entry` GRanges must carry these `mcols`: `SNP`, `A1` (effect /
dosage-counted allele), `A2` (other allele), `Z`, `N`; optionally `MAF`,
`INFO`, `BETA`, `SE`, `P`. Two settings apply to the whole collection and are
stored in slots, not columns:

* `genome` — the build label (e.g. `"GRCh38"`).
* `ldSketch` — a `GenotypeHandle` giving the LD reference used for QC and downstream analyses.

Optional per-row columns: `varY` (phenotype variance) and, for case/control
GWAS, `nCase` / `nControl`.

`QtlSumStats` and `GwasSumStats` can be constructed in two ways:

* **From a manifest** — `loadGwasSumStatsFromManifest()` /
  `loadQtlSumStatsFromManifest()` read a table pointing at the sumstats files
  and assemble the collection.
* **Manually** — build the `entry` GRanges yourself and call the
  `GwasSumStats()` / `QtlSumStats()` constructor.

> **The loaders run no QC.** They build the raw collection with an empty
> `qcInfo` slot. Allele harmonization, MAF/INFO/N filtering, and LD-mismatch QC
> are a separate, mandatory step — `summaryStatsQc()` — documented in the
> *Quality Control for GWAS and QTL Summary Statistics* vignette. Downstream
> pipelines reject collections whose `qcInfo` is empty.

```{r load-packages}
library(pecotmr)
```

## Toy data for this vignette

We will use the `toy_ref` PLINK1 panel (chr22) as the LD reference, and a handful of
its variants as our summary statistics so the LD-overlap check succeeds.

```{r toy-data}
genoPrefix <- sub("\\.bed$", "",
                  system.file("extdata", "toy_ref.bed", package = "pecotmr"))
ldSketch <- GenotypeHandle(plink1Prefix = genoPrefix)

# Five real toy_ref chr22 sites (positions and alleles taken from the panel).
sumstatsDf <- data.frame(
  chrom      = rep("22", 5),
  pos        = c(14560203, 14564328, 14850625, 14870204, 14878387),
  variant_id = c("rs11089128", "rs7288972", "rs11167319", "rs8138488", "rs2186521"),
  A1         = c("G", "C", "G", "C", "A"),
  A2         = c("A", "T", "T", "T", "G"),
  z          = c(1.20, -0.42, 2.13, 0.05, -1.83),
  n_sample   = rep(10000L, 5),
  stringsAsFactors = FALSE)

workdir <- file.path(tempdir(), "sumstats-vignette")
dir.create(workdir, showWarnings = FALSE)
gwasTsv <- file.path(workdir, "study1.tsv")
readr::write_tsv(sumstatsDf, gwasTsv)
```

## GwasSumStats from a manifest

The GWAS manifest has **one row per study**. `genome` and `ldSketch` are
collection-level, so they are supplied as arguments (or as constant
`genome` / `ldSketchPath` columns).

| Column | Required | Meaning |
|---|---|---|
| `study` (alias `study_id`) | yes, unique | study identifier |
| `sumStatsPath` (aliases `path`, `file_path`) | yes | sumstats file |
| `columnMapping` (alias `column_mapping`) | no | YAML column map (see below) |
| `nCase` / `nControl` | no | case / control counts |
| `varY` | no | phenotype variance |
| `genome`, `ldSketchPath` | single-valued | may instead be arguments |

```{r gwas-from-manifest}
gwasManifest <- data.frame(
  study        = "MyGWAS",
  sumStatsPath = gwasTsv,
  stringsAsFactors = FALSE)

gws <- loadGwasSumStatsFromManifest(
  gwasManifest, genome = "GRCh38", ldSketch = ldSketch)
gws
S4Vectors::mcols(gws$entry[[1]])
```

The `ldSketch` argument also accepts a path/prefix or a chromosome-sharded
`genoMeta` specification (see `?GenotypeHandle`), so a genome-wide reference can
be supplied as a `#chr,path` meta file rather than a single handle.

## GwasSumStats manually

The loader is a wrapper over the `GwasSumStats()` constructor. Built by hand,
the pattern is: make one `GRanges` per study with the required `mcols`, then
call the constructor with the shared `genome` and `ldSketch`.

```{r gwas-manual}
entry <- GenomicRanges::GRanges(
  seqnames = paste0("chr", sumstatsDf$chrom),
  ranges   = IRanges::IRanges(start = sumstatsDf$pos, width = 1L))
S4Vectors::mcols(entry) <- S4Vectors::DataFrame(
  SNP = sumstatsDf$variant_id,
  A1  = sumstatsDf$A1, A2 = sumstatsDf$A2,
  Z   = sumstatsDf$z,  N  = sumstatsDf$n_sample)

gwsManual <- GwasSumStats(
  study    = "MyGWAS",
  entry    = list(entry),
  genome   = "GRCh38",
  ldSketch = ldSketch)
gwsManual
```

For a case/control study, pass `nCase` / `nControl`; for sufficient statistics interface with continuous traits, pass `varY`.

## QtlSumStats from a manifest

The QTL manifest has **one row per `(study, context, trait)` tuple**. Each row
points at a sumstats file for that trait.

| Column | Required | Meaning |
|---|---|---|
| `study` | yes | study identifier |
| `context` (alias `cond`) | yes | context / condition |
| `trait` (alias `gene_id`) | yes | molecular trait id |
| `sumStatsPath` (alias `path`) | yes | sumstats file for this trait |
| `columnMapping`, `varY` | no | as for GWAS |
| `genome`, `ldSketchPath` | single-valued | may instead be arguments |

```{r qtl-from-manifest}
qtlManifest <- data.frame(
  study        = "MyQTL",
  context      = "Blood",
  trait        = "ENSG00000001",
  sumStatsPath = gwasTsv,           # reuse the toy table for one gene
  stringsAsFactors = FALSE)

qts <- loadQtlSumStatsFromManifest(
  qtlManifest, genome = "GRCh38", ldSketch = ldSketch)
qts
```

## QtlSumStats manually

```{r qtl-manual}
qtsManual <- QtlSumStats(
  study    = "MyQTL",
  context  = "Blood",
  trait    = "ENSG00000001",
  entry    = list(entry),
  genome   = "GRCh38",
  ldSketch = ldSketch)
qtsManual
```

For a multi-context collection (e.g. as input to `mash`), pass parallel
`study` / `context` / `trait` vectors and a matching list of `entry` GRanges —
one element per tuple.

## Reading from files

### Column mapping (YAML)

When columns in a  sumstats file don't match the standard names, supply a
`columnMapping` — a YAML file of `standardName: sourceName` entries (the
xqtl-protocol format), either per-row in the manifest or as a default argument.
Standard keys are `chrom`, `pos`, `variant_id`, `A1`, `A2`, `z`, `n_sample`
(and optional `beta`, `se`, `p`, `maf`, `info`).

```{r column-mapping}
oddDf <- sumstatsDf
names(oddDf) <- c("CHR", "POS", "RSID", "EA", "OA", "ZSCORE", "SAMPLE_N")
oddTsv <- file.path(workdir, "odd_columns.tsv")
readr::write_tsv(oddDf, oddTsv)

mapPath <- file.path(workdir, "column_map.yaml")
yaml::write_yaml(list(chrom = "CHR", pos = "POS", variant_id = "RSID",
                      A1 = "EA", A2 = "OA", z = "ZSCORE",
                      n_sample = "SAMPLE_N"), mapPath)

gwsMapped <- loadGwasSumStatsFromManifest(
  data.frame(study = "MyGWAS", sumStatsPath = oddTsv, columnMapping = mapPath),
  genome = "GRCh38", ldSketch = ldSketch)
length(gwsMapped$entry[[1]])
```

### Region-restricted reads (tabix)

If a `region` is supplied and the sumstats file is **compressed with bgzip and
indexed with tabix** (a `<file>.tbi` must exist in the same directory), only that region is read.
Plain-text files are read whole and the region is ignored with a warning. The
tabix header line must begin with `#` so its column names can be recovered.

```{r tabix-region}
tabixDf <- sumstatsDf
names(tabixDf)[1] <- "#chrom"      # tabix header convention
tabixTsv <- file.path(workdir, "study1.sorted.tsv")
readr::write_tsv(tabixDf, tabixTsv)
gz <- Rsamtools::bgzip(tabixTsv, dest = paste0(tabixTsv, ".gz"), overwrite = TRUE)
Rsamtools::indexTabix(gz, seq = 1L, start = 2L, end = 2L, comment = "#")

gwsRegion <- loadGwasSumStatsFromManifest(
  data.frame(study = "MyGWAS", sumStatsPath = gz),
  genome = "GRCh38", ldSketch = ldSketch,
  region = "22:14560000-14565000")
length(gwsRegion$entry[[1]])   # only the two variants in the region
```

### GWAS-VCF

VCF summary statistics are read as **GWAS-VCF**: the effect allele `A1` is the
`ALT` allele, `A2` is `REF`, and the statistics come from the per-study FORMAT
fields `ES`, `SE`, `LP`, `SS`, `EAF` (`Z = ES / SE`, `N = SS`, `P = 10^-LP`,
`MAF = fold(EAF)`). Reading only a specific requires the VCF to be compressed with bgzip and indexed with tabix. **BCF is
not supported** — convert to a bgzipped VCF first. Support for reading VCFs requires the
`VariantAnnotation` package.

```{r gwas-vcf, eval=hasVA}
vcfLines <- c(
  "##fileformat=VCFv4.2",
  '##FORMAT=<ID=ES,Number=A,Type=Float,Description="effect size">',
  '##FORMAT=<ID=SE,Number=A,Type=Float,Description="standard error">',
  '##FORMAT=<ID=LP,Number=A,Type=Float,Description="-log10 p">',
  '##FORMAT=<ID=SS,Number=A,Type=Float,Description="sample size">',
  "##contig=<ID=22>",
  paste("#CHROM", "POS", "ID", "REF", "ALT", "QUAL", "FILTER", "INFO",
        "FORMAT", "MyGWAS", sep = "\t"))
for (i in seq_len(nrow(sumstatsDf))) {
  vcfLines <- c(vcfLines, paste(
    "22", sumstatsDf$pos[i], sumstatsDf$variant_id[i],
    sumstatsDf$A2[i], sumstatsDf$A1[i],   # REF = A2, ALT = A1
    ".", "PASS", ".", "ES:SE:LP:SS",
    sprintf("%.3f:0.100:1.000:10000", sumstatsDf$z[i] * 0.1), sep = "\t"))
}
vcfPath <- file.path(workdir, "study1.vcf")
writeLines(vcfLines, vcfPath)

gwsVcf <- loadGwasSumStatsFromManifest(
  data.frame(study = "MyGWAS", sumStatsPath = vcfPath),
  genome = "GRCh38", ldSketch = ldSketch)
S4Vectors::mcols(gwsVcf$entry[[1]])[, c("SNP", "A1", "A2", "Z", "N")]
```

For a multi-sample GWAS-VCF, pass `sampleSelect = "<column>"` to pick the
study; override the FORMAT tag names with `formatMapping` if they differ from
`ES`/`SE`/`LP`/`SS`/`EAF`.

## Next steps

The collections carry no QC yet. Run `summaryStatsQc()` before any downstream
analysis (it populates `qcInfo`; downstream pipelines require it):

```{r qc, eval=FALSE}
gws_qcd <- summaryStatsQc(gws, zMismatchQc = "slalom")
```

See the *Quality Control for GWAS and QTL Summary Statistics* vignette for the
full QC pipeline. `QtlSumStats` and `GwasSumStats` are direct inputs to
`fineMappingPipeline()`, `twasWeightsPipeline()` (`QtlSumStats` only) and `colocBoostPipeline()`— see the *Fine-mapping with pecotmr*, *Learning TWAS weights with pecotmr* and *Multi-trait colocalization with ColocBoost* vignettes.

```{r accessors}
getGenome(gws)
as.character(gws$study)
length(getSumStats(gws))   # variants in the first entry
```

```{r session-info}
sessionInfo()
```
