---
title: "Building QtlDataset and MultiStudyQtlDataset objects"
author: "pecotmr authors"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Constructing QtlDataset and MultiStudyQtlDataset objects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5)
```

## Overview

A `QtlDataset` is an individual-level container for a **single study**: one
genotype source plus a per-context set of molecular phenotypes.
`MultiStudyQtlDataset` bundles several `QtlDataset`s (optionally alongside a
summary-statistics-only `QtlSumStats`) so a molecular trait measured across multiple studies can
be analyzed with downstream pipelines together.

| Slot | Holds |
|---|---|
| `study` | study identifier (length-1 character) |
| `genotypes` | a `GenotypeHandle` (lazy PLINK1/PLINK2/VCF/GDS handle) |
| `phenotypes` | a **named list of `SummarizedExperiment`s**, one per context; assay = traits x samples, `rowRanges` = trait positions, `colData` = per-context covariates |
| `genotypeCovariates` | numeric matrix (samples x covariates), e.g. genotype PCs, applied to every context |
| QC thresholds | `mafCutoff`, `macCutoff`, `xvarCutoff`, `imissCutoff`, `keepIndel`, `keepSamples`, `keepVariants`, `scaleResiduals` — stored as **lazy** filters, applied at genotype-extraction time |

There are two ways to build a `QtlDataset`:

* **From a manifest** — `loadQtlDatasetFromManifest()` reads a table (`data.frame`
  or file) that points at the phenotype / covariate / genotype files and
  assembles the object for you. This is the recommended route for real data.
* **Manually** — build the per-context `SummarizedExperiment`s and the
  `GenotypeHandle` yourself and call the `QtlDataset()` constructor. Useful when
  your data is already in memory.

```{r load-packages}
library(pecotmr)
```

## Toy data for this vignette

We will reuse the `toy_ref` PLINK1 panel shipped with the package as the genotype
source, take its first few samples as our phenotyped individuals, and write two
small per-context phenotype BED files (TensorQTL-style layout: `#chr`, `start`, `end`,
`gene_id`, then one column per sample) plus a covariate table to a temporary
directory.

```{r toy-data}
genoPrefix <- sub("\\.bed$", "",
                  system.file("extdata", "toy_ref.bed", package = "pecotmr"))
samples <- head(getSampleIds(GenotypeHandle(plink1Prefix = genoPrefix)), 40)

workdir <- file.path(tempdir(), "qtl-vignette")
dir.create(workdir, showWarnings = FALSE)

# Two genes on chr22, positioned inside the toy_ref variant span so a cis
# window captures nearby variants.
writePhenoBed <- function(path, seed) {
  set.seed(seed)
  bed <- data.frame(
    `#chr`  = c("chr22", "chr22"),
    start   = c(14600000L, 14900000L),
    end     = c(14600001L, 14900001L),
    gene_id = c("GENE1", "GENE2"),
    check.names = FALSE)
  for (s in samples) bed[[s]] <- round(rnorm(2), 3)
  readr::write_tsv(bed, path)
  path
}
muscleBed <- writePhenoBed(file.path(workdir, "Muscle.bed"), 1)
bloodBed  <- writePhenoBed(file.path(workdir, "Blood.bed"),  2)

# Per-context covariate table: samples as rows, covariate columns after an ID
# column (here two genotype PCs).
covPath <- file.path(workdir, "covariates.tsv")
set.seed(3)
readr::write_tsv(
  data.frame(sample = samples,
             PC1 = round(rnorm(length(samples)), 3),
             PC2 = round(rnorm(length(samples)), 3),
             check.names = FALSE),
  covPath)
```

## Constructing a QtlDataset from a manifest

The QtlDataset manifest has **one row per context**. Canonical column names
are camelCase; snake_case aliases such as `cond`, `cov_path`, `genotype_prefix`
are also accepted and normalized on read.

| Column | Required | Meaning |
|---|---|---|
| `context` (alias `cond`) | yes | context / condition label |
| `phenotypePath` (alias `path`) | yes | bgzipped or plain BED phenotype file |
| `covariatePath` (alias `cov_path`) | no | per-context covariate table |
| `study` | single-valued | study identifier (may also be an argument) |
| `genotypePath` (alias `genotype_prefix`) | single-valued | genotype file or PLINK prefix (may also be an argument) |
| `genotypeCovariatePath` | no, single-valued | genotype-derived covariates applied to all contexts |

`study` and `genotypePath` are single-valued for the whole dataset, so you can
supply them either as a constant manifest column **or** as an argument; if you
give both, they must agree.

```{r qtl-from-manifest}
manifest <- data.frame(
  context       = c("Muscle", "Blood"),
  phenotypePath = c(muscleBed, bloodBed),
  covariatePath = c(covPath, covPath),
  study         = "GTEx_toy",
  genotypePath  = genoPrefix,
  stringsAsFactors = FALSE)

qd <- loadQtlDatasetFromManifest(manifest)
qd
```

The manifest can equally be a path to a `.csv` (read as CSV) or any other
extension (read as TSV):

```{r qtl-manifest-file}
manifestPath <- file.path(workdir, "qtl_manifest.tsv")
qd <- loadQtlDatasetFromManifest(manifestPath)
```

The QC thresholds are ordinary arguments (they are stored as lazy filters, not
applied at load), for example:

```{r qtl-manifest-qc}
qdFiltered <- loadQtlDatasetFromManifest(
  manifest, mafCutoff = 0.01, imissCutoff = 0.05, keepIndel = FALSE)
```

## Constructing a QtlDataset manually

The manifest loader is a thin wrapper over the `QtlDataset()` constructor. Doing
it by hand shows exactly what the object contains: a named list of
`SummarizedExperiment`s and a `GenotypeHandle`.

```{r qtl-manual}
# One SummarizedExperiment per context. Rows are traits, columns are samples;
# rowRanges carries each trait's position, colData carries per-context
# covariates.
buildSE <- function(bedPath) {
  bed <- as.data.frame(readr::read_tsv(bedPath, show_col_types = FALSE),
                       check.names = FALSE)
  sampleCols <- setdiff(names(bed), c("#chr", "start", "end", "gene_id"))
  expr <- as.matrix(bed[, sampleCols, drop = FALSE])
  rownames(expr) <- bed$gene_id
  rr <- GenomicRanges::GRanges(
    seqnames = bed$`#chr`,
    ranges   = IRanges::IRanges(start = bed$start + 1L, end = bed$end))
  names(rr) <- bed$gene_id
  cov <- as.data.frame(readr::read_tsv(covPath, show_col_types = FALSE),
                       check.names = FALSE)
  cd <- S4Vectors::DataFrame(cov[match(sampleCols, cov$sample),
                                 c("PC1", "PC2"), drop = FALSE],
                             row.names = sampleCols)
  SummarizedExperiment::SummarizedExperiment(
    assays = list(expression = expr), rowRanges = rr, colData = cd)
}

phenotypes <- list(Muscle = buildSE(muscleBed), Blood = buildSE(bloodBed))
genotypes  <- GenotypeHandle(plink1Prefix = genoPrefix)

qdManual <- QtlDataset(
  study      = "GTEx_toy",
  genotypes  = genotypes,
  phenotypes = phenotypes)
qdManual
```

This approach gives users more flexibility if they need to make more modifications to the molecular phenotype data after loading it from a file but before making a `QtlDataset`

## MultiStudyQtlDataset

A `MultiStudyQtlDataset` needs at least two studies in total - at least two individual-level studies or at least one individual-level study and one summary-only study. When loading from a manifest,
a **`qtlDatasetsManifest`** with `study` and `genotypePath` must be provided where rows 
are grouped by `study` and one `QtlDataset` is built per group. A `sumStatsManifest` describing summary-only studies can also be provided (see the *Constructing
QtlSumStats and GwasSumStats objects* vignette).

```{r multistudy-manifest}
multiManifest <- data.frame(
  study         = c("StudyA", "StudyB"),
  context       = c("Muscle", "Muscle"),
  phenotypePath = c(muscleBed, bloodBed),
  genotypePath  = genoPrefix,
  stringsAsFactors = FALSE)

msd <- loadMultiStudyQtlDatasetFromManifest(multiManifest)
msd
```

A MultiStudyQtlDataset can be manually constructed by building a named list of `QtlDataset`s and passing it to the constructor:

```{r multistudy-manual}
msdManual <- MultiStudyQtlDataset(
  qtlDatasets = list(StudyA = qd, StudyB = qdManual))
names(msdManual@qtlDatasets)
```

To attach a summary-only study, pass a `QtlSumStats` object to the `sumStats` argument.
manifest form, supply `sumStatsManifest`, `genome`, and `ldSketch`).

## Next steps

Once built, the accessors extract data lazily, applying the stored QC filters
at extraction time:

```{r accessors}
getStudy(qd)
getContexts(qd)
# Cis genotypes around a trait (+/- a window), QC applied on the fly:
dim(getGenotypes(qd, traitId = "GENE1", cisWindow = 1e6))
# Phenotype matrix for one context:
dim(SummarizedExperiment::assay(getPhenotypes(qd, contexts = "Muscle")))
```

`QtlDataset` and `MultiStudyQtlDataset` are direct inputs to
`fineMappingPipeline()`, `twasWeightsPipeline()` and `colocBoostPipeline()`— see the *Fine-mapping with pecotmr*, *Learning TWAS weights* and *Multi-trait colocalization with ColocBoost* vignettes.

```{r session-info}
sessionInfo()
```
