---
title: "Building QtlDataset and MultiStudyQtlDataset objects"
author: "pecotmr authors"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Building QtlDataset and MultiStudyQtlDataset objects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    message = FALSE,
    warning = FALSE,
    fig.width = 7,
    fig.height = 4.5
)
```

## Overview

A `QtlDataset` is an individual-level container for a **single study**: one
genotype source plus a per-context set of molecular phenotypes.
`MultiStudyQtlDataset` bundles several `QtlDataset`s (optionally alongside a
summary-statistics-only `QtlSumStats`) so a molecular trait measured across
multiple studies can be analyzed with downstream pipelines together.

It **is** a [`MultiAssayExperiment`][mae], so the whole multi-assay surface
works on it directly. Each QTL context is one experiment, and the genotypes
are one more:

[mae]: https://bioconductor.org/packages/MultiAssayExperiment/

*  **`genotype`** — `rowRanges` is one range per variant (carrying `A1` /
    `A2`); the `dosage` assay is variants x samples, read **lazily** from the
    genotype files; `colData` holds genotype-derived covariates such as
    ancestry PCs.
*  **one per context** — `rowRanges` is trait positions, the assay is
    traits x samples, and `colData` holds that context's phenotype
    covariates.

Alongside them the object keeps a `colData` naming every sample and a
`sampleMap` recording which of them each experiment observes — so contexts
need not share a sample set — plus its own slots:

*  `study` — study identifier (length-1 character).
*  `genotypes` — the genotype panel (a lazy PLINK1/PLINK2/VCF/GDS source,
    read with `readGenotypes()`) that the dosage assay reads through.
*  QC thresholds — `mafCutoff`, `macCutoff`, `xvarCutoff`, `imissCutoff`,
    `keepIndel`, `keepVariants`, `scaleResiduals`. Stored as **lazy** filters,
    applied at genotype-extraction time.

To restrict a dataset to a set of samples, subset it: `x[, samples, ]` narrows
`colData` and `sampleMap` together, and keeps the class and its slots. The
`keepSamples` constructor argument is shorthand for the same thing.

There are two ways to build a `QtlDataset`:

*  **From a manifest** — `loadQtlDatasetFromManifest()` reads a table
    (`data.frame` or file) that points at the phenotype / covariate / genotype
    files and assembles the object for you. This is the recommended route for
    real data.
*  **Manually** — build the per-context `SummarizedExperiment`s and the
    genotype panel yourself and call the `QtlDataset()` constructor. Useful
    when your data is already in memory.

```{r load-packages}
library(pecotmr)
```

## Toy data for this vignette

We will reuse the `toy_ref` PLINK1 panel shipped with the package as the
genotype source, take its first few samples as our phenotyped individuals, and
write two small per-context phenotype BED files (TensorQTL-style layout:
`#chr`, `start`, `end`, `gene_id`, then one column per sample) plus a covariate
table to a temporary directory.

```{r toy-data}
genoPrefix <- sub(
    "\\.bed$",
    "",
    system.file("extdata", "toy_ref.bed", package = "pecotmr")
)
samples <- head(colnames(readGenotypes(plink1Prefix = genoPrefix)), 40)

workdir <- file.path(tempdir(), "qtl-vignette")
dir.create(workdir, showWarnings = FALSE)

# Two genes on chr22, positioned inside the toy_ref variant span so a cis
# window captures nearby variants.
writePhenoBed <- function(path, seed) {
    set.seed(seed)
    bed <- data.frame(
        `#chr` = c("chr22", "chr22"),
        start = c(14600000L, 14900000L),
        end = c(14600001L, 14900001L),
        gene_id = c("GENE1", "GENE2"),
        check.names = FALSE
    )
    for (s in samples) {
        bed[[s]] <- round(rnorm(2), 3)
    }
    readr::write_tsv(bed, path)
    path
}
muscleBed <- writePhenoBed(file.path(workdir, "Muscle.bed"), 1)
bloodBed <- writePhenoBed(file.path(workdir, "Blood.bed"), 2)

# Per-context covariate table: samples as rows, covariate columns after an ID
# column (here two genotype PCs).
covPath <- file.path(workdir, "covariates.tsv")
set.seed(3)
readr::write_tsv(
    data.frame(
        sample = samples,
        PC1 = round(rnorm(length(samples)), 3),
        PC2 = round(rnorm(length(samples)), 3),
        check.names = FALSE
    ),
    covPath
)
```

## Constructing a QtlDataset from a manifest

The QtlDataset manifest has **one row per context**. Canonical column names
are camelCase; snake_case aliases such as `cond`, `cov_path`, `genotype_prefix`
are also accepted and normalized on read.

| Column | Required | Meaning |
|---|---|---|
| `context` (alias `cond`) | yes | context / condition label |
| `phenotypePath` (alias `path`) | yes | bgzipped or plain BED phenotype file |
| `covariatePath` (alias `cov_path`) | no | per-context covariate table |
| `study` | single-valued | study identifier (may also be an argument) |
| `genotypePath` (alias `genotype_prefix`) | single-valued | genotype file or PLINK prefix (may also be an argument) |
| `genotypeCovariatePath` | no, single-valued | genotype-derived covariates applied to all contexts |

`study` and `genotypePath` are single-valued for the whole dataset, so you can
supply them either as a constant manifest column **or** as an argument; if you
give both, they must agree.

```{r qtl-from-manifest}
manifest <- data.frame(
    context = c("Muscle", "Blood"),
    phenotypePath = c(muscleBed, bloodBed),
    covariatePath = c(covPath, covPath),
    study = "GTEx_toy",
    genotypePath = genoPrefix,
    stringsAsFactors = FALSE
)

# The path columns hold absolute paths; show basenames for readability
# (the real `manifest` keeps the full paths the loader needs).
transform(
    manifest,
    phenotypePath = basename(phenotypePath),
    covariatePath = basename(covariatePath),
    genotypePath = basename(genotypePath)
)

qd <- loadQtlDatasetFromManifest(manifest)
qd
```

The manifest can equally be a path to a `.csv` (read as CSV) or any other
extension (read as TSV):

```{r qtl-manifest-file}
manifestPath <- file.path(workdir, "qtl_manifest.tsv")
readr::write_tsv(manifest, manifestPath)
qd <- loadQtlDatasetFromManifest(manifestPath)
```

The QC thresholds are ordinary arguments (they are stored as lazy filters, not
applied at load), for example:

```{r qtl-manifest-qc}
qdFiltered <- loadQtlDatasetFromManifest(
    manifest,
    mafCutoff = 0.01,
    imissCutoff = 0.05,
    keepIndel = FALSE
)
```

## Constructing a QtlDataset manually

The manifest loader is a thin wrapper over the `QtlDataset()` constructor. Doing
it by hand shows exactly what the object contains: a named list of
`SummarizedExperiment`s and a genotype panel.

```{r qtl-manual}
# One SummarizedExperiment per context. Rows are traits, columns are samples;
# rowRanges carries each trait's position, colData carries per-context
# covariates.
buildSE <- function(bedPath) {
    bed <- as.data.frame(
        readr::read_tsv(bedPath, show_col_types = FALSE),
        check.names = FALSE
    )
    sampleCols <- setdiff(names(bed), c("#chr", "start", "end", "gene_id"))
    expr <- as.matrix(bed[, sampleCols, drop = FALSE])
    rownames(expr) <- bed$gene_id
    rr <- GenomicRanges::GRanges(
        seqnames = bed$`#chr`,
        ranges = IRanges::IRanges(start = bed$start + 1L, end = bed$end)
    )
    names(rr) <- bed$gene_id
    cov <- as.data.frame(
        readr::read_tsv(covPath, show_col_types = FALSE),
        check.names = FALSE
    )
    cd <- S4Vectors::DataFrame(
        cov[match(sampleCols, cov$sample), c("PC1", "PC2"), drop = FALSE],
        row.names = sampleCols
    )
    SummarizedExperiment::SummarizedExperiment(
        assays = list(expression = expr),
        rowRanges = rr,
        colData = cd
    )
}

phenotypes <- list(Muscle = buildSE(muscleBed), Blood = buildSE(bloodBed))
genotypes <- readGenotypes(plink1Prefix = genoPrefix)

qdManual <- QtlDataset(
    study = "GTEx_toy",
    genotypes = genotypes,
    phenotypes = phenotypes
)
qdManual
```

This approach gives users more flexibility if they need to make more
modifications to the molecular phenotype data after loading it from a file but
before making a `QtlDataset`

## MultiStudyQtlDataset

A `MultiStudyQtlDataset` needs at least two studies in total - at least two
individual-level studies or at least one individual-level study and one
summary-only study. When loading from a manifest, a **`qtlDatasetsManifest`**
with `study` and `genotypePath` must be provided where rows are grouped by
`study` and one `QtlDataset` is built per group. A `sumStatsManifest`
describing summary-only studies can also be provided (see the *Constructing
QtlSumStats and GwasSumStats objects* vignette).

```{r multistudy-manifest}
multiManifest <- data.frame(
    study = c("StudyA", "StudyB"),
    context = c("Muscle", "Muscle"),
    phenotypePath = c(muscleBed, bloodBed),
    genotypePath = genoPrefix,
    stringsAsFactors = FALSE
)

msd <- loadMultiStudyQtlDatasetFromManifest(multiManifest)
msd
```

A MultiStudyQtlDataset can be manually constructed by building a named list of
`QtlDataset`s and passing it to the constructor:

```{r multistudy-manual}
msdManual <- MultiStudyQtlDataset(
    qtlDatasets = list(StudyA = qd, StudyB = qdManual)
)
names(getQtlDatasets(msdManual))
```

To attach a summary-only study, pass a `QtlSumStats` object to the `sumStats`
argument. manifest form, supply `sumStatsManifest`, `genome`, and `ldSketch`).

## Excluding related samples

Both `fineMappingPipeline()` and `twasWeightsPipeline()` fit models that assume
independent samples. Related individuals break that assumption: they inflate
the effective sample size, so cross-validated R-squared comes back optimistic
and PIPs overconfident. Relatedness is therefore a QC step applied *before* the
dataset is built — the pipelines do not model it.

`filterRelatedness()` takes a kinship table — one row per pair, the shape
PLINK's `--genome` or KING emits — and returns the individuals to **drop**:

```{r relatedness, eval = requireNamespace("plinkQC", quietly = TRUE) && requireNamespace("igraph", quietly = TRUE)}
relatedness <- data.frame(
    IID1 = c(samples[1], samples[3], samples[5]),
    IID2 = c(samples[2], samples[4], samples[6]),
    PI_HAT = c(0.50, 0.25, 0.125),
    stringsAsFactors = FALSE
)
excluded <- filterRelatedness(relatedness, relatednessThreshold = 0.0625)
excluded
unrelated <- setdiff(samples, excluded)
length(unrelated)
```

One member of each related pair comes back, chosen so that what remains is as
large an unrelated set as possible. `relatednessThreshold` defaults to 0.0625,
which is third-degree relatives; raise it to 0.125 to keep everyone but
second-degree and closer.

The exclusion list is not applied for you — feed it to `keepSamples`, the
constructor argument this vignette introduced above, or subset an existing
dataset with `qd[, unrelated, ]`. Both narrow `colData` and `sampleMap`
together.

`analysisType` changes what is being maximised. The default,
`"maximizeUnrelated"`, keeps the largest unrelated set. `"maximizeCases"`
instead retains as many cases as it can — the right objective for a
case/control design, where cases are the scarce resource — and requires
`phenoData` naming each sample's status.

`filterRelatedness()` needs the suggested packages **igraph** and **plinkQC**;
the chunk above is skipped when they are not installed.

## Next steps

Once built, the accessors extract data lazily, applying the stored QC filters
at extraction time:

```{r accessors}
getStudy(qd)
getContexts(qd)
# Cis genotypes around a trait (+/- a window), QC applied on the fly:
dim(getGenotypes(qd, traitId = "GENE1", cisWindow = 1e6))
# Phenotype matrix for one context:
dim(SummarizedExperiment::assay(getPhenotypes(qd, contexts = "Muscle")))
```

`QtlDataset` and `MultiStudyQtlDataset` are direct inputs to
`fineMappingPipeline()`, `twasWeightsPipeline()` and `colocboostPipeline()`—
see the *Fine-mapping with pecotmr*, *Learning TWAS weights with pecotmr* and
*Multi-trait colocalization with ColocBoost* vignettes.

```{r session-info}
sessionInfo()
```
