---
title: "Causal TWAS with ctwasPipeline"
author: "pecotmr authors"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Causal TWAS with ctwasPipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
hasCtwas <- requireNamespace("ctwas", quietly = TRUE)
knitr::opts_chunk$set(
    echo = TRUE,
    message = FALSE,
    warning = FALSE,
    fig.width = 7,
    fig.height = 4.5,
    eval = hasCtwas
)
# cTWAS narrates each step through the logging package, which writes to
# stdout rather than as messages, so knitr cannot filter it. Quieten it here
# to keep the output readable -- in your own session the log is worth
# watching, since it reports region counts and EM convergence.
if (hasCtwas) {
    logging::setLevel("WARN")
}
```

```{r ctwas-missing, echo=FALSE, eval=!hasCtwas, results="asis"}
cat(
    "**Note:** the `ctwas` package is not installed, so the code below is",
    "shown but not run. Install it from",
    "<https://github.com/xinhe-lab/multigroup_ctwas>."
)
```

## Overview

A TWAS Z-score (see the
[TWAS Z-score + MR vignette](twas-zscore.html)) asks whether a gene's
predicted expression associates with the trait. It asks that question one
gene at a time, so a gene can score highly purely because it sits near a
strong GWAS signal that is really driven by a nearby variant.

`ctwasPipeline()` runs
[cTWAS](https://github.com/xinhe-lab/multigroup_ctwas), which fine-maps
**genes and SNPs jointly within each LD block**. A gene's posterior
inclusion probability is therefore conditioned on the SNP signal beside it,
which is what separates a gene that carries the association from one that
merely neighbours it.

Inputs:

| Argument | What it is |
|---|---|
| `gwasSumStats` | one `GwasSumStats` whose elements are LD blocks |
| `twasWeights` | per-gene `TwasWeights` / `QtlFineMappingResult` |
| `twasZ` | optional pre-computed per-gene Z from `causalInferencePipeline()` |
| `fineMappingResult` | optional; enables the credible-set and PIP variant rescues |

```{r load-packages}
library(pecotmr)
```

## The LD block convention

`gwasSumStats` is **one** `GwasSumStats` object whose elements are LD
blocks, keyed by its `blockId` column — not a list of per-block objects.
Build it with `loadGwasSumStatsFromManifest(..., ldBlocks = <blocks>)` and
pass it through `summaryStatsQc()`.

A single-block input is rejected: cTWAS estimates its group priors by EM
across regions, and one region gives it nothing to average over.

```{r bundled}
ldStem <- file.path(
    system.file("extdata", "ld_reference", "chr22", package = "pecotmr"),
    "protocol_example.LD.chr22"
)
gwasTsv <- system.file(
    "extdata",
    "manifests",
    "protocol_example.twas.gwas_sumstats.chr22.tsv.gz",
    package = "pecotmr"
)

blocks <- GenomicRanges::GRanges(
    "chr22",
    IRanges::IRanges(c(10000000, 15000001), c(15000000, 19000000)),
    blockId = c("chr22_1", "chr22_2")
)

gwasSumStats <- loadGwasSumStatsFromManifest(
    manifest = data.frame(study = "gwas1", sumStatsPath = gwasTsv),
    genome = "hg38",
    ldSketch = ldStem,
    region = "chr22:10000000-19000000",
    ldBlocks = blocks
)
gwasByRegion <- summaryStatsQc(gwasSumStats, mafCutoff = 0.0025)

data(ctwasWeightsExample)
```

## Running the pipeline

```{r ctwas-run}
res <- ctwasPipeline(
    gwasSumStats = gwasByRegion,
    twasWeights = list(ctwasWeightsExample),
    thin = 1,
    niterPrefit = 3,
    niter = 10,
    min_group_size = 1,
    min_p_single_effect = 0,
    min_nonSNP_PIP = 0,
    fallbackToPrefit = TRUE
)
res
```

The settings above are sized for a toy dataset and are **not** defaults you
should carry into a real analysis:

- `thin = 1` keeps every SNP (production runs thin to `0.1`).
- `niterPrefit` / `niter` are cut to keep the vignette fast.
- `min_nonSNP_PIP = 0` disables region screening. The bundled GWAS has no
  genome-wide-significant signal, so the default (`0.5`) selects no region at
  all and the result comes back empty. On real data, leave it alone.
- `fallbackToPrefit = TRUE` recovers from the accurate-EM step returning
  `NaN` on underpowered data by keeping the prefit estimates.

## Where genes come from

`twasWeights` accepts two shapes:

- a **flat** `TwasWeights` or `QtlFineMappingResult` (or a homogeneous list
  of them) carrying `region` provenance. Each gene is assigned to its home
  LD block by `start(region)`, matching cTWAS's own placement rule.
- a **pre-bucketed named list** keyed by `region_id`, used as-is.

Blocks with no gene weights still contribute their SNP-level signal to the
group prior — the whole-chromosome pattern where only a few blocks carry
genes.

A gene whose cis window straddles a block boundary is assigned by its single
anchor. The signal it loses across the boundary is a separate concern,
handled by `mergeCtwasBoundaryRegions()` below.

## Trimming each gene's variants

Weight matrices from a permissive QTL model can be large, and cTWAS's cost
scales with them. Four arguments trim each gene before cTWAS sees it:

| Argument | Effect |
|---|---|
| `twasWeightCutoff` | drop variants with `abs(weight)` below the cutoff |
| `maxNumVariants` | cap variants per gene |
| `csMinCor` | keep variants in a credible set with purity at least this |
| `minPipCutoff` | keep variants with PIP above this |

`csMinCor` and `minPipCutoff` mark variants as must-keep, so they survive
the `maxNumVariants` cap. Both need a `fineMappingResult`; without one they
are no-ops and the cap falls back to ordering by `abs(weight)`.

```{r ctwas-filters, eval=FALSE}
ctwasPipeline(
    gwasSumStats = gwasByRegion,
    twasWeights = list(ctwasWeightsExample),
    fineMappingResult = fmr,
    twasWeightCutoff = 1e-4,
    maxNumVariants = 1000,
    csMinCor = 0.8,
    minPipCutoff = 0.1
)
```

## Reading the result

`ctwasPipeline()` returns a `CtwasResult`: one row per
`(gwasStudy, study, context, method)` run.

```{r ctwas-identity}
getMethodNames(res)
getStudy(res)
getContexts(res)
```

The per-gene posteriors live in `getFinemap()`:

```{r ctwas-finemap}
fm <- getFinemap(res)
as.data.frame(fm)[, c(
    "molecular_id",
    "context",
    "region_id",
    "z",
    "susie_pip",
    "cs"
)]
```

`susie_pip` is the quantity cTWAS exists to produce: the probability the
gene is causal given the SNPs it competes with. Here it is low (the toy GWAS
has no real signal) and `cs` is `NA` because the gene falls in no credible
set.

`getSusieAlpha()` returns the per-single-effect alpha matrix behind those
PIPs, and the EM group priors sit on each row's entry:

```{r ctwas-param}
param <- getCtwasParam(res$entry[[1]])
param$group_prior
param$group_prior_var
param$converged
```

By default the context-agnostic SNP background is dropped from the result,
since it is the null the genes are being scored against rather than a
finding. Pass `keepSnps = TRUE` to retain it as an extra row
(`study = context = "SNP"`), which makes the full cTWAS output
reconstructable.

## The four chainable steps

`ctwasPipeline()` is a wrapper over four steps. Call them directly when you
want to inspect an intermediate payload, override the estimated priors, or
re-enter the chain without recomputing what came before.

```{r ctwas-granular}
inputs <- assembleCtwasInputs(
    gwasSumStats = gwasByRegion,
    twasWeights = list(ctwasWeightsExample)
)
est <- estCtwasParam(
    inputs,
    thin = 1,
    niterPrefit = 3,
    niter = 10,
    min_group_size = 1,
    min_p_single_effect = 0,
    fallbackToPrefit = TRUE
)
screened <- screenCtwasRegions(est, min_nonSNP_PIP = 0)
finemapped <- finemapCtwasRegions(screened)
nrow(finemapped$finemap_res)
```

`finemap_res` here has more rows than `getFinemap(res)` above: it still
carries the SNP rows. `asCtwasResult()` is the step that structures the
payload and drops them.

```{r ctwas-as-result}
nrow(getFinemap(asCtwasResult(finemapped)))
nrow(getFinemap(asCtwasResult(finemapped, keepSnps = TRUE)))
```

Each step's payload is also bundled, so you can enter anywhere:

```{r ctwas-bundled-payloads, eval=FALSE}
data(ctwasInputsExample) # after assembleCtwasInputs
data(ctwasEstExample) # after estCtwasParam
data(ctwasFinemapExample) # after finemapCtwasRegions
```

The most common reason to use the granular path is to override the priors.
`estCtwasParam()`'s accurate-EM step can diverge on underpowered data;
`fallbackToPrefit = TRUE` handles that automatically, but taking the steps
apart lets you substitute priors of your own into `screenCtwasRegions()`.

## Boundary genes

A gene whose cis window crosses an LD block boundary is fine-mapped in only
one of the two blocks. `mergeCtwasBoundaryRegions()` re-runs such genes over
the merged pair:

```{r ctwas-merge, eval=FALSE}
merged <- mergeCtwasBoundaryRegions(
    finemapped,
    pipThresh = 0.5,
    filterCs = FALSE,
    maxSNP = Inf
)
```

`ctwasPipeline(mergeBoundary = TRUE)` does the same thing inline, with
`mergePipThresh` / `mergeFilterCs` / `mergeMaxSNP` forwarding to those
arguments. It is off by default because it only matters when a
high-PIP gene actually straddles a boundary.

## Common parameters

```{r ctwas-params, eval=FALSE}
ctwasPipeline(
    gwasSumStats = gwasByRegion,
    twasWeights = list(ctwasWeightsExample),
    method = NULL, # NULL: "ensemble" if present, else every method
    thin = 0.1,
    niterPrefit = 3,
    niter = 30,
    L = 5,
    groupPriorVarStructure = "shared_type",
    ncore = 1,
    keepSnps = FALSE,
    mergeBoundary = FALSE,
    fallbackToPrefit = FALSE
)
```

When `method` is `NULL` and several weight methods are present without an
`"ensemble"`, each method runs as its own independent cTWAS run and
contributes its own row-set.

## Next steps

- [TWAS weights](twas-weights.html) — training the weights cTWAS consumes.
- [TWAS Z-scores and MR](twas-zscore.html) — the per-gene Z that
  `twasZ` optionally supplies here.
- [QC for summary statistics](rss-qc.html) — `summaryStatsQc()`, which
  `gwasSumStats` must have been through.

# Session info

```{r session-info}
sessionInfo()
```
