---
title: "Fine-mapping with pecotmr"
author: "pecotmr authors"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fine-mapping with pecotmr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    message = FALSE,
    warning = FALSE,
    fig.width = 7,
    fig.height = 4.5
)
```

## Overview

`fineMappingPipeline()` is a single pipeline that can accept any of the
following inputs:

| Input class | What it represents | Methods supported |
|---|---|---|
| `QtlDataset` | Single-study individual-level data | susie, susieInf, susieAsh, mvsusie, fsusie, susieSer |
| `MultiStudyQtlDataset` | Multiple `QtlDataset` objects (optionally with an embedded `QtlSumStats`) | susie, susieInf, susieAsh, mvsusie, fsusie (individual-level only), susieSer (individual-level only)|
| `QtlSumStats` | QTL summary statistics annotated by `(study, context, trait)` | susie, susieInf, susieAsh, mvsusie |
| `GwasSumStats` | GWAS summary statistics annotated by `study` | susie, susieInf, susieAsh, mvsusie |

Method arguments are the same across input classes; `fineMappingPipeline()`
routes `methods = "susie"` to `susieR::susie` for individual-level data
and to `susieR::susie_rss` for summary-statistics inputs automatically.

The output is always a `FineMappingResult` (a `GRangesList`-subclass
collection, one element per fine-mapped tuple).

```{r load-packages}
library(pecotmr)
```

## Bundled inputs

```{r bundled-inputs}
data(
    qtlDatasetExample,
    qtlSumStatsExample,
    gwasSumStatsS4Example,
    multiStudyQtlDatasetExample
)

qtlDatasetExample
```

The `*_sumstats_example` objects ship with a non-empty `qcInfo` slot so
they bypass the `summaryStatsQc()` gate. In a real analysis you'd
construct the raw `QtlSumStats` / `GwasSumStats` from your data, then
run `summaryStatsQc()` to populate that slot (see the
[summary-statistics QC vignette](rss-qc.html)).

## Individual-level fine-mapping

`fineMappingPipeline()` always needs to know which genotype window to
fit on. With an individual-level `QtlDataset`, pass `cisWindow` (in
basepairs around each trait's TSS), supply an explicit `region`, or
set both per-call.

```{r individual-fm}
fmr <- fineMappingPipeline(
    qtlDatasetExample,
    methods = "susie",
    cisWindow = 1e6
)
fmr
```

Per-tuple PIPs:

```{r individual-pips}
pip <- getPip(
    fmr,
    study = "study1",
    context = "brain",
    trait = "ENSG_example",
    method = "susie"
)
head(sort(pip, decreasing = TRUE), 5)
```

Credible sets:

```{r individual-cs}
getCs(
    fmr,
    study = "study1",
    context = "brain",
    trait = "ENSG_example",
    method = "susie",
    coverage = 0.95
)
```

The `top_loci` table consolidates per-method PIPs, CS membership, and
sumstats into one long-format data.frame:

```{r individual-toploci}
head(
    getTopLoci(
        fmr,
        study = "study1",
        context = "brain",
        trait = "ENSG_example",
        method = "susie"
    ),
    5
)
```

## Multi-study fine-mapping

A `MultiStudyQtlDataset` runs the pipeline per-study and
concatenates the results into a single `QtlFineMappingResult`:

```{r multistudy-fm}
msFmr <- fineMappingPipeline(
    multiStudyQtlDatasetExample,
    methods = "susie",
    cisWindow = 1e6
)
table(msFmr$study)
```

## QTL summary-statistics fine-mapping

```{r sumstats-qtl}
fmrSs <- fineMappingPipeline(qtlSumStatsExample, methods = "susie")
fmrSs
```

The `QtlSumStats` collection carries an `ldSketch` genotype panel; the
pipeline pulls the per-region LD from there automatically.

## GWAS summary-statistics fine-mapping

```{r sumstats-gwas}
fmrGwas <- fineMappingPipeline(gwasSumStatsS4Example, methods = "susie")
fmrGwas
```

The GWAS path is similar to the QTL-sumstats path but only annotates by study.

## Common parameters

```{r fm-params, eval=FALSE}
fineMappingPipeline(
    qtlDatasetExample,
    methods = c("susie", "susieInf"),
    contexts = "brain",
    traitId = "ENSG_example",
    region = "chr22:25000000-26000000",
    cisWindow = 5e5,
    coverage = 0.95,
    secondaryCoverage = c(0.7, 0.5),
    signalCutoff = 0.025,
    minAbsCorr = 0.8,
    addSusieInf = TRUE,
    naAction = "drop", # "drop" (default) or "impute"
    verbose = 1
)
```

-  `naAction` controls how phenotype NAs are handled by
    `getResidualizedPhenotypes()`: `"drop"` removes samples with any NA across
    the requested traits, `"impute"` mean-imputes each trait independently.
-  `addSusieInf = TRUE` runs SuSiE-inf first and uses its result to initialise
    the susie fit, which improves credible-set purity in regions with
    background polygenic signal.
-  `secondaryCoverage` adds secondary CS columns at the requested coverages
    (alongside the primary `coverage`); useful when you want both narrow and
    broad CS reports.

## Joint multi-context / multi-trait fits

`jointSpecification` triggers the dispatch for cross-axis joint
methods (`mvsusie` for cross-context, `mvsusie` or `fsusie` for multi-trait):

```{r fm-joint, eval=FALSE}
fineMappingPipeline(
    multiStudyQtlDatasetExample,
    methods = "mvsusie",
    jointSpecification = list(axis = "context", contexts = c("brain", "liver"))
)
```

See `?jointSpecification` for the specification grammar.

## Reading the result

`FineMappingResult` is a `GRangesList` subclass, one element per
`(study, context, trait, method)` tuple. The element is the fitted variant
set as a `GRanges`, so the result is addressable by genomic range as well as
by tuple; the tuple itself and the per-tuple payloads live in `mcols()`,
reachable with `$`:

```{r fm-shape}
fmr
colnames(fmr)
lengths(fmr) # variants per tuple
```

Rather than reaching into the payload, use the view accessors. Each returns a
tidy table over the whole collection, with the tuple columns carried along:

```{r fm-views}
head(getTopLoci(fmr), 3) # per-variant posterior summary
head(getCs(fmr), 3) # credible-set membership
head(getCredibleSetSummary(fmr), 3) # one row per credible set
```

Passing a tuple selects a single fit and returns that tuple's values on their
own — a bare vector rather than a table:

```{r fm-one-tuple}
key <- list(
    study = fmr$study[[1L]],
    context = fmr$context[[1L]],
    trait = fmr$trait[[1L]],
    method = fmr$method[[1L]]
)
pip <- do.call(getPip, c(list(fmr), key))
length(pip)
```

The underlying SuSiE fit is available with `getSusieFit()` and the
cross-validation record with `getCvResult()`, both keyed the same way.
Range-based subsetting works directly:

```{r fm-range}
window <- GenomicRanges::GRanges(
    "chr22",
    IRanges::IRanges(20000000, 21000000)
)
sum(lengths(subsetRegion(fmr, window)))
```

## Next steps

-  `QtlFineMappingResult` is used by [`colocPipeline`](coloc-pipeline.html) for
    QTL-GWAS colocalization and by [`twasWeightsPipeline`](twas-weights.html)
    so that fine-mapping weights are contributed to the TWAS ensemble weights
    model alongside regularized regression methods.
-  Run the full [`causalInferencePipeline`](twas-zscore.html) (TWAS + MR) with
    a `FineMappingResult` and `GwasSumStats`.

# Session info

```{r session-info}
sessionInfo()
```
