---
title: "Post-processing fine-mapping results"
author: "pecotmr authors"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Post-processing fine-mapping results}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    message = FALSE,
    warning = FALSE,
    fig.width = 7,
    fig.height = 4.5
)
```

## Overview

[`fineMappingPipeline()`](fine-mapping.html) returns a
`QtlFineMappingResult` or `GwasFineMappingResult`. This vignette covers what
you do with one afterwards: summarising credible sets, comparing results
across studies, intersecting QTL with GWAS, and reaching the underlying
SuSiE quantities when the summaries are not enough.

Every function here reads an existing result — none of them refit anything.

```{r load-packages}
library(pecotmr)
data(qtlFineMappingExample, gwasFineMappingExample)
data(qtlFineMappingLbfExample, gwasFineMappingLbfExample)
data(qtlFineMappingPairedExample, qtlDatasetExample)
qtlFineMappingExample
```

## The two tables

A fine-mapping result carries two flat summaries, and which one you want
depends on whether your unit of interest is a variant or a credible set.

**`getTopLoci()`** is one row per retained variant, with its PIP, effect
size, and credible-set membership at each coverage level:

```{r top-loci}
topLoci <- getTopLoci(qtlFineMappingExample)
dim(topLoci)
head(as.data.frame(topLoci)[, c(
    "variant_id",
    "pip",
    "cs_95",
    "cs_95_purity",
    "beta",
    "se"
)], 4)
```

`cs_95` names the credible set the variant belongs to, or is `NA` when it
belongs to none. The `cs_70` / `cs_50` columns give the same membership at
looser coverage, so a variant can be in a 95% set but not a 50% one.

**`getCredibleSetSummary()`** is one row per credible set instead:

```{r cs-summary}
as.data.frame(getCredibleSetSummary(qtlFineMappingExample))[, c(
    "cs",
    "coverage",
    "n_variants",
    "purity_min",
    "cs_log10bf",
    "lead_variant",
    "lead_pip"
)]
```

`purity_min` is the smallest absolute correlation between any pair of
variants in the set — the standard measure of whether a credible set is a
single signal or an artefact of a few uncorrelated variants being lumped
together. `cs_log10bf` says how much the data prefer that effect being
present.

A credible set with low `purity_min` deserves suspicion before
interpretation, which is why the pipeline exposes `minPurity` as a filter
rather than dropping such sets silently.

## Merging overlapping credible sets

A multi-effect SuSiE fit can return credible sets that overlap, especially
when two effects are in strong LD. `mergeSusieCs()` collapses those into one
row per variant, keeping the smallest credible-set index a variant belongs
to:

```{r merge-cs}
merged <- mergeSusieCs(qtlFineMappingExample, coverage = 0.95)
dim(merged)
head(as.data.frame(merged)[, c("variant_id", "cs", "pip")], 4)
```

Use it when downstream code needs a variant to belong to at most one set —
overlap-counting, for instance, would otherwise double-count.

## Intersecting QTL and GWAS

`overlapTopLoci()` joins two fine-mapping results on shared variants,
prefixing each side's columns. It is the cheap first pass before
[colocalization](coloc-pipeline.html): if the two credible sets share no
variants at all, there is nothing for a coloc model to work with.

```{r overlap}
overlap <- overlapTopLoci(qtlFineMappingLbfExample, gwasFineMappingLbfExample)
nrow(overlap)
head(as.data.frame(overlap)[, c(
    "variant_id",
    "qtl_pip",
    "gwas_pip",
    "qtl_cs_95",
    "gwas_cs_95"
)], 4)
```

The variants worth attention are those in a credible set on *both* sides:

```{r overlap-both}
both <- overlap[!is.na(overlap$qtl_cs_95) & !is.na(overlap$gwas_cs_95), ]
nrow(both)
table(both$qtl_cs_95, both$gwas_cs_95)
```

Here that filter removes nothing, because `topLoci` in this result retains
only credible-set variants — every shared variant is in one set on each
side. The filter earns its place on results where `topLoci` also carries
high-PIP variants outside any set, which is the usual case once
`pipCutoff` is in play.

A non-empty overlap is not evidence of a shared causal variant, however
large: two independent signals in the same LD block overlap too. That
question is what `colocPipeline()` answers.

## Combining results

`combineFineMappingResults()` stacks several results into one collection,
which is how per-region or per-chromosome runs are assembled after
parallel execution:

```{r combine}
combined <- combineFineMappingResults(
    qtlFineMappingExample,
    qtlFineMappingLbfExample
)
nrow(combined)
as.data.frame(combined)[, c("study", "context", "trait", "method")]
```

The inputs keep their own identity columns, so a combined object still
distinguishes which study and method each row came from.

## Between-credible-set correlation

`computeCsCorrelation()` reports how correlated a fit's credible sets are
with each other. It is computed on demand rather than stored, because it
needs LD — and the LD lives with the data, not with the fit:

```{r cs-corr}
computeCsCorrelation(qtlFineMappingPairedExample, qtlDatasetExample)
```

The `ldSource` argument takes a `QtlDataset`, `QtlSumStats` or
`GwasSumStats` — whichever object the fit was derived from. It must cover
the fit's region: a source whose variants do not reach the fit's window
cannot supply the correlations, and the function says so rather than
returning something empty.

Near-zero off-diagonal entries, as here, mean the credible sets are
capturing distinct signals. Strongly correlated sets suggest the fit split
one signal in two.

## Reaching the SuSiE quantities

When the flat tables are not enough — you want the per-effect posterior, or
to re-derive PIPs under a different assumption — the underlying fit is
available.

`getLbf()` returns the log Bayes factors, one row per variant and one
column per single effect:

```{r lbf}
lbf <- getLbf(qtlFineMappingLbfExample)
unit <- lbf[lbf$trait == lbf$trait[1] & lbf$context == lbf$context[1], ]
lbfMatrix <- as.matrix(unit[, grep("^lbf_L", names(unit))])
dim(lbfMatrix)
```

`lbfToAlpha()` converts those to single-effect inclusion probabilities.
It normalises **across each row**, so the matrix has to be oriented with
effects as rows — transpose the variants-by-effects layout above:

```{r lbf-alpha}
alpha <- lbfToAlpha(t(lbfMatrix))
dim(alpha)
rowSums(alpha) # each single effect sums to 1 across variants
```

Getting that orientation wrong is quiet rather than loud: normalising across
effects for each variant also produces a matrix whose rows sum to 1, just
not one that means anything.

The LBF matrix is also what `colocPipeline()` consumes, which is why
`fineMappingPipeline(..., keepLbf = TRUE)` is required before
colocalization.

`getSusieFit()` returns the raw per-unit SuSiE object when you need
something the accessors do not expose:

```{r susie-fit}
fit <- getSusieFit(qtlFineMappingExample)
class(fit)
names(fit)
```

`alpha` and `lbf_variable` are the per-effect matrices, `sets` the credible
sets before any purity filtering, and `V` the estimated prior variances —
everything the flat tables above are derived from.

## How the tables are built

The summaries above are not stored by the fitter — they are derived from the
raw SuSiE fit by a short chain that `fineMappingPipeline()` runs for you:

```
postprocessFinemappingFits()   # per-method fits -> a post object
  -> computeCsTables()         # credible sets from alpha, with purity
  -> buildTopLoci()            # the flat one-row-per-variant table
  -> formatFinemappingOutput() # assemble the result payload
```

Call them directly when you have SuSiE fits from elsewhere and want
pecotmr's tables over them, or when you need to change how credible sets are
formed — `computeCsTables()` is where the coverage levels and purity
threshold are applied, so re-running it with a different `coverage` gives a
different `top_loci` without refitting anything.

Two smaller extractors sit at the row level and are what
`getCredibleSetSummary()` is assembled from:

- `extractCsInfo(fmRow, csNames, topLociTable, ldSource)` — per-credible-set
  detail for one result row, including the between-variant correlations that
  give `purity_min`, which is why it needs an `ldSource`.
- `extractTopPipInfo(fmRow, sumstats)` — the lead variant and its statistics,
  joined against summary statistics.

`getSusieResult()` unwraps a per-condition fit list into the single SuSiE
object the other helpers expect.

## Writing results out

`writeSumStatsVcf()` serialises a fine-mapping result — or summary
statistics — to GWAS-VCF, the interchange format most external tools read:

```{r vcf, eval=FALSE}
writeSumStatsVcf(
    qtlFineMappingExample,
    file = "finemapping.vcf.gz",
    genome = "hg38"
)
```

It needs the `VariantAnnotation` package, which is a suggested dependency
rather than a hard one.

## Next steps

- [Fine-mapping](fine-mapping.html) — producing the results this vignette
  consumes.
- [Colocalization](coloc-pipeline.html) — the model that answers whether an
  overlap reflects a shared causal variant.
- [TWAS weights](twas-weights.html) — reusing a fine-mapping fit as a weight
  method.

# Session info

```{r session-info}
sessionInfo()
```
