---
title: "SNP heritability and functional enrichment"
author: "pecotmr authors"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SNP heritability and functional enrichment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    message = FALSE,
    warning = FALSE,
    fig.width = 7,
    fig.height = 4.5
)
```

## Overview

`estimateH2()` estimates SNP heritability from GWAS summary statistics and
an LD reference. Four estimators are available, differing in what they model
and what they need:

| `method` | Approach | LD input |
|---|---|---|
| `"sldsc"` | Stratified LD score regression (Finucane 2015) | `LdScore` |
| `"gldsc"` | Generalized LDSC — GLS with full block LD (Xiong 2024) | `LdScore` + matrices |
| `"lder"` | LD eigenvalue regression (Song 2022) | `LdEigen` |
| `"hdl"` | High-definition likelihood (Ning 2020) | `LdEigen` |

Supplying `annotations` makes any of them stratified, adding a
per-annotation enrichment table.

There is a second, separate track in the package: when stratified LDSC has
already been run externally by [polyfun][pf], `sldscPostprocessingPipeline()`
postprocesses those outputs. Both are covered below.

[pf]: https://github.com/omerwe/polyfun

```{r load-packages}
library(pecotmr)
data(ldScoreExample, ldEigenExample)
ldScoreExample
```

Both bundled references describe the same 1300 variants across 14 chr22 LD
blocks, taken from the de-identified `protocol_example` panel, so the four
methods can be compared on one dataset.

## Variants must line up

Every estimator reads the z-scores positionally, indexed by the LD
reference's own variant order — nothing is matched on variant ID along the
way. `estimateH2()` therefore checks up front that the summary statistics,
the annotations and the reference describe the same variants, in the same
order, on the same genome build, and refuses to run otherwise.

The practical consequence: build the summary statistics *from* the reference
rather than alongside it.

```{r simulate-gwas}
# A GWAS simulated on this reference so the true h2 is known: z is drawn
# from the standard model z ~ N(0, N R diag(v) R + R), block by block.
M <- length(ldScoreExample)
nGwas <- 10000
h2True <- 0.4
perSnpVar <- rep(h2True / M, M)

set.seed(1)
z <- numeric(M)
for (block in getLdMatrixList(ldScoreExample)) {
    idx <- block$snpIdx
    p <- length(idx)
    sigma <- nGwas * (block$R %*% diag(perSnpVar[idx], p) %*% block$R) +
        block$R
    e <- eigen((sigma + t(sigma)) / 2, symmetric = TRUE)
    z[idx] <- as.vector(e$vectors %*% (sqrt(pmax(e$values, 0)) * rnorm(p)))
}

gr <- as(ldScoreExample, "GRanges")
S4Vectors::mcols(gr) <- S4Vectors::DataFrame(
    SNP = names(ldScoreExample),
    A1 = S4Vectors::mcols(ldScoreExample)$A1,
    A2 = S4Vectors::mcols(ldScoreExample)$A2,
    Z = z,
    N = as.integer(nGwas)
)
gwas <- GwasSumStats(
    study = "trait1",
    entry = list(gr),
    genome = getGenome(ldScoreExample),
    ldSketch = readGenotypes(
        system.file("extdata", "toy_ref.bed", package = "pecotmr")
    )
)
```

## Estimating heritability

```{r estimate}
methods <- c("sldsc", "gldsc", "lder", "hdl")
estimates <- lapply(methods, function(m) {
    ref <- if (m %in% c("sldsc", "gldsc")) ldScoreExample else ldEigenExample
    suppressWarnings(estimateH2(gwas, ref, method = m))
})
names(estimates) <- methods

data.frame(
    method = methods,
    h2 = round(vapply(estimates, getH2, numeric(1)), 3),
    se = round(vapply(estimates, getH2Se, numeric(1)), 3),
    row.names = NULL
)
```

The true value is `r h2True`. Three of the four land close to it; `sldsc`
comes in low, and that is worth understanding rather than tuning away.

**`sldsc` is attenuated because the reference is small.** LD score
regression is a genome-scale method — real analyses regress over roughly a
million variants. On 1300 the intercept is not identifiable: an unweighted
OLS on this same data returns an intercept several times its true value of
1, and the heritability absorbs the difference. `lder` is the least
sensitive to reference size, which is why it is the default.

All four are checked against their reference implementations — [ldsc][ldsc],
[HDL][hdl], [LDER][lder] and [gldsc][gldsc] — on identical inputs: HDL agrees
exactly, LDER to machine precision, gLDSC to ~3e-8 and S-LDSC to ~1e-6. The
small residuals are a ridge on the weighted solve, which keeps an exactly
singular design from returning `NA`. Agreement holds even where the
reference implementations misbehave, such as an S-LDSC intercept below 1 or
a negative partitioned heritability.

One deliberate difference: LDER here clamps negative eigenvalues before
forming its regression weights. A real, rank-deficient LD block decomposes
with a tail of tiny negative values that are floating-point noise, and
upstream takes `sqrt()` of the resulting negative weight and aborts. The
clamp changes nothing wherever upstream runs.

[ldsc]: https://github.com/bulik/ldsc
[hdl]: https://github.com/zhenin/HDL
[lder]: https://github.com/shuangsong0110/LDER
[gldsc]: https://github.com/xzw20046/gldsc

**`hdl` has a reference-size limit, and it tells you when it hits it.** It
is the only one of the four that models reference-panel *noise* explicitly:
its likelihood subtracts a term scaling as `1/nRef` from the modelled
variance. Once the reference is small relative to the GWAS, that term
outgrows the signal, the modelled variance is driven to its floor across
most eigen-directions, and the optimiser walks the intercept out to the top
of its range — returning a number that reflects that bound rather than the
data. At `N = 1e5` against this reference (`nRef` =
`r getNRef(ldEigenExample)`) it drifts to about 0.56 for a true 0.4 and
warns that the fit sits on a bound. At the `N = 1e4` used here it is
comfortably inside its range. If HDL disagrees sharply with the others,
check for that warning and compare `getN()` against `getNRef()`.

A standard error of zero, or an estimate pinned at a boundary, almost always
means the reference is too small rather than that the trait has no
heritability. The block jackknife needs at least two LD blocks and wants many
more; `estimateH2()` refuses a single-block reference outright.

## Building an LD reference

The bundled references were built with `buildLdEigen()` / `buildLdScore()`,
which turn already-loaded LD into the statistic each estimator needs. Load
the LD **per block** — loading a whole region at once returns one dense
matrix, which is one block, and one block has no jackknife.

```{r build-ref, eval=FALSE}
metaPath <- "/path/to/ld_meta_file.tsv"
blocks <- read.delim(metaPath, check.names = FALSE)
regions <- sprintf("chr22:%d-%d", blocks$start, blocks$end)

ldList <- lapply(regions, function(r) loadLdMatrix(metaPath, region = r))

ldEigen <- buildLdEigen(ldList, genome = "hg38")
ldScore <- buildLdScore(ldList, genome = "hg38")
```

`nRef` and the genome build default to what the loaded LD records.
`buildLdScore(keepLdMatrices = FALSE)` drops the per-block matrices, which
shrinks the object by roughly an order of magnitude — but only unstratified
`"sldsc"` can use the result, since `"gldsc"` and stratified `"sldsc"` both
form their per-annotation scores from those matrices.

## Stratified estimation and enrichment

Passing an `AnnotationMatrix` partitions heritability across annotations.
The annotation rows must line up with the reference, exactly as the summary
statistics do.

```{r annotate}
# An illustrative annotation over the first half of the reference, simulated
# to carry three times the per-variant heritability of the rest.
inA <- seq_len(M) <= M / 2
perSnpVarStrat <- ifelse(inA, 0.225 / sum(inA), 0.075 / sum(!inA))

set.seed(1)
zStrat <- numeric(M)
for (block in getLdMatrixList(ldScoreExample)) {
    idx <- block$snpIdx
    p <- length(idx)
    sigma <- nGwas *
        (block$R %*% diag(perSnpVarStrat[idx], p) %*% block$R) +
        block$R
    e <- eigen((sigma + t(sigma)) / 2, symmetric = TRUE)
    zStrat[idx] <- as.vector(
        e$vectors %*% (sqrt(pmax(e$values, 0)) * rnorm(p))
    )
}
S4Vectors::mcols(gr)$Z <- zStrat
gwasStrat <- GwasSumStats(
    study = "trait1",
    entry = list(gr),
    genome = getGenome(ldScoreExample),
    ldSketch = readGenotypes(
        system.file("extdata", "toy_ref.bed", package = "pecotmr")
    )
)

annot <- AnnotationMatrix(
    matrix(as.integer(inA), ncol = 1, dimnames = list(NULL, "annotA")),
    as(ldScoreExample, "GRanges"),
    data.frame(name = "annotA", tier = "baseline", type = "binary"),
    genome = getGenome(ldScoreExample)
)
```

```{r enrichment}
strat <- estimateH2(
    gwasStrat,
    ldScoreExample,
    method = "gldsc",
    annotations = annot
)
as.data.frame(getEnrichment(strat))[, c(
    "annotation",
    "tau",
    "enrichment",
    "enrichmentSe",
    "propH2",
    "propSnps"
)]
```

`enrichment` is the proportion of heritability divided by the proportion of
variants: above 1 means the annotation carries more heritability than its
share of the genome. The simulated truth here is 3, and the estimate is
directionally right but attenuated — again a consequence of the reference's
size rather than of the annotation.

Annotations come from files via `readAnnotations()`, which reads BED, BigWig
and LDSC `.annot` formats at a given set of variant positions:

```{r read-annotations, eval=FALSE}
annot <- readAnnotations(
    c(enhancer = "enhancers.bed", conservation = "phastCons.bw"),
    snpRanges = as(ldScoreExample, "GRanges"),
    genome = "hg38"
)
```

The `tier` column decides how each annotation is used: `"baseline"`
annotations enter the model and get an enrichment estimate, while
`"candidate"` annotations are scored against the fitted baseline (the sHDL
score test, via `getScoreStats()`).

One structural caveat: a set of baseline annotations that exactly partitions
the variants is rank-deficient by construction, because the model's implicit
all-ones base annotation is then their sum. Real baseline models — the LDSC
baselineLD set, for instance — use overlapping annotations and do not have
this problem.

## Postprocessing external polyfun runs

The second track handles stratified LDSC that was run outside R by polyfun.
The readers load its outputs, `SldscData` bundles them, and the pipeline
does the computation — no file I/O of its own.

```{r sldsc-read}
sldscDir <- system.file("extdata", "sldsc", package = "pecotmr")
traitDir <- system.file("extdata", "sldsc_trait", package = "pecotmr")

annotTable <- readSldscAnnot(sldscDir)
frq <- readSldscFrq(sldscDir, plinkName = "reference.")

traitNames <- c(traitA = "sumstats", traitB = "sumstats2", traitC = "sumstats3")
runs <- lapply(traitNames, function(t) {
    readSldscTrait(file.path(traitDir, paste0(t, ".parquet")))
})
runs$traitA$categories
```

Those categories are a real baseline model — `Coding_UCSC`,
`Conserved_LindbladToh` and `Repressed_Hoffman`, each split into common and
low-frequency, plus `base` — with `ANNOT_0` as the target annotation under
test.

```{r sldsc-pipeline}
sldscData <- SldscData(
    annot = annotTable,
    frq = frq,
    traits = lapply(runs, function(r) list(single = list(r)))
)
result <- sldscPostprocessingPipeline(
    sldscData,
    targetCategories = "ANNOT_0"
)
as.data.frame(result$meta$enrichment)
```

The pipeline standardizes each trait's tau to Gazal tau\*, back-solves a
jackknife standard error for the enrichment statistic, and pools traits by
DerSimonian-Laird random-effects meta-analysis. `sldscSubsetMeta()` re-runs
the meta over a subset of traits without refitting anything:

```{r sldsc-subset}
subset <- sldscSubsetMeta(result, subsetTraits = c("traitB", "traitC"))
unlist(subset$enrichment$ANNOT_0)
```

Each polyfun run may be a `single` target run or a `joint` one; supplying
both lets the pipeline report the target's enrichment from the single fit
alongside its coefficient from the joint fit.

### The postprocessing steps

`sldscPostprocessingPipeline()` is a wrapper over steps that are exported
individually, which is what you want when auditing a surprising number:

```{r sldsc-steps}
computeSldscMRef(sldscData, mafCutoff = 0.05) # the M_ref that scales tau*
isBinarySldscAnnot(sldscData) # binary vs continuous
```

`computeSldscAnnotSd()` gives the per-annotation standard deviation that
turns tau into Gazal tau\*, `standardizeSldscTrait()` does that
standardisation for one trait's run, and `metaSldscRandom()` is the
DerSimonian-Laird pooling across traits. Running them yourself is the way to
see which of the three a strange tau\* came from.

`getAnnotData()`, `getFrqData()`, `getAnnotCols()` and `getTraitRuns()` read
the corresponding parts of an `SldscData` back out.

### LD scores from an existing reference

`computeLdScores()` derives per-variant LD scores from an `LdEigen` or
`LdScore`, optionally stratified by annotation:

```{r compute-ld-scores}
scores <- computeLdScores(ldEigenExample)
head(round(as.vector(scores[, 1]), 3))
```

This is the same quantity `buildLdScore()` stores, reconstructed from the
eigendecomposition — `sum_i V[j,i]^2 d[i]^2` rather than `sum_k r^2_jk`.
The two agree, which is worth knowing when you have one object and need the
other's input.

## Bridging the two tracks

`h2EstimateToSldscTrait()` converts an `H2Estimate` into the same list shape
`readSldscTrait()` produces, so an in-package estimate can be meta-analysed
alongside polyfun runs:

```{r bridge}
names(h2EstimateToSldscTrait(strat))
```

## Next steps

- [Molecular QTL enrichment](xqtl-enrichment.html) — the complementary
  per-variant QTL/GWAS enrichment analysis.
- [QC for summary statistics](rss-qc.html) — preparing the GWAS side.

# Session info

```{r session-info}
sessionInfo()
```
