---
title: "Working with LD references"
author: "pecotmr authors"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with LD references}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    message = FALSE,
    warning = FALSE,
    fig.width = 7,
    fig.height = 4.5
)
```

## Overview

Almost everything else in pecotmr needs LD. Fine-mapping from summary
statistics needs it, RSS weight learning needs it, colocalization needs it,
and the heritability estimators are built on nothing else. The pipelines
handle it for you, so this vignette is about the layer underneath: loading
LD for a region, computing it from genotypes, checking it is usable, and
reducing a variant set that is too correlated to model.

```{r load-packages}
library(pecotmr)
ldMeta <- system.file(
    "extdata",
    "ld_reference",
    "ld_meta_file.tsv",
    package = "pecotmr"
)
```

An LD reference is described by a **meta file**: one row per LD block, giving
the block's coordinates and the path to the genotype or correlation data
backing it.

```{r meta}
read.delim(ldMeta, check.names = FALSE)
```

Paths are relative to the meta file, so a reference directory can be moved
without rewriting it.

## Loading LD for a region

`loadLdMatrix()` returns an `LdData` — a `GRanges` of the variants plus the
correlation matrix, or the genotypes to compute it from:

```{r load}
region <- "chr22:10000000-19000000"
ld <- loadLdMatrix(ldMeta, region = region)
ld
R <- getCorrelation(ld)
dim(R)
```

Whether a correlation matrix or a genotype handle comes back depends on the
source. `returnGenotype = "auto"` (the default for genotype sources) returns
dosages and computes `R` on demand, which matters because a dense matrix for
a large region is far bigger than the genotypes it came from.

`getVariantInfo()` gives the variants as a plain `GRanges`, which is what you
need to align summary statistics against the panel:

```{r variant-info}
head(getVariantInfo(ld), 3)
```

**A region is not a block.** Loading a whole region returns one dense matrix
covering it. To get one matrix per LD block — which the block jackknife in
[heritability estimation](heritability.html) requires — address the blocks
individually.

`region` also takes a vector, returning one `LdData` per entry — which is
how you get per-block LD without writing the loop yourself:

```{r multi-region}
blocks <- read.delim(ldMeta, check.names = FALSE)
ldList <- loadLdMatrix(
    ldMeta,
    region = sprintf("chr22:%d-%d", blocks$start, blocks$end)
)
length(ldList)
```

### Sources without coordinates

An LD meta file carries coordinates, so its blocks are addressed by `region`.
Other sources have no coordinates and are addressed by `block` index instead:

- an **`ldInfo` table** — an `LD_file` column and optionally `SNP_file`, the
  shape cTWAS's `LD_map` uses
- a **matrix** already in memory, or a list of them

```{r by-block}
R2 <- computeLd(matrix(rnorm(100 * 10), 100, 10))
loadLdMatrix(R2, block = 1)
```

Supplying both `region` and `block`, or neither, is an error — the source
determines which one applies.

### Options

| Argument | Effect |
|---|---|
| `returnGenotype` | `"auto"` gives dosages for genotype sources, `R` for pre-computed |
| `dropMonomorphic` | drop variants with no variation |
| `materializeGenotypes` | read dosages once and keep them, so later access does no file I/O |
| `maxVariants` / `seed` | randomly thin oversized blocks, reproducibly |

```{r options}
thinned <- loadLdMatrix(ldMeta, region = region, maxVariants = 50, seed = 1)
length(thinned)
```

## Computing LD from genotypes

`computeLd()` takes a dosage matrix and returns the correlation matrix:

```{r compute}
set.seed(1)
X <- matrix(rnorm(100 * 20), 100, 20, dimnames = list(NULL, paste0("v", 1:20)))
X[, 5] <- X[, 4] + rnorm(100, sd = 0.01) # a near-duplicate pair
R2 <- computeLd(X)
dim(R2)
```

`method` selects the estimator — `"sample"` (default), `"population"`, or
`"gcta"` — and `shrinkage` applies shrinkage toward the identity. `backend`
can hand the work to SNPRelate or snpStats when those are installed.

## Checking that LD is usable

A correlation matrix estimated from finite samples is usually not positive
definite, and methods that invert it or take its eigendecomposition will
fail or return nonsense. `checkLd()` reports the diagnosis. Run it on the
real panel rather than the toy matrix above, which is well conditioned by
construction:

```{r check}
diagnosis <- checkLd(R)
diagnosis[c(
    "isPd",
    "isPsd",
    "minEigenvalue",
    "nNegative",
    "conditionNumber"
)]
```

That is what a real reference looks like: positive *semi*-definite but not
positive definite, with a smallest eigenvalue at rounding scale (`-7e-16`)
and a condition number around `1e9`. The negative value is floating-point
noise on a rank-deficient matrix, not structure — but it is exactly what
makes an unguarded `sqrt()` or `solve()` fail, and why LDER clamps
eigenvalues before forming its weights.

`method = "shrink"` or `"eigenfix"` returns a repaired matrix instead of a
report:

```{r check-fix}
fixed <- checkLd(R, method = "eigenfix")
c(before = diagnosis$minEigenvalue, after = checkLd(fixed$R)$minEigenvalue)
```

The smallest eigenvalue is lifted off zero, though on a matrix this
rank-deficient it stays tiny — repair makes the decomposition safe, it does
not restore rank that the reference panel never had. Repairing also changes
the data, so prefer a better-conditioned variant set where you can, which is
what the next section is for.

## Reducing a correlated variant set

Two variants in near-perfect LD carry the same information, and a design
containing both is singular. Three tools address that, differing in what
they optimise.

`ldPruneByCorrelation()` drops variants until no pair exceeds a correlation
threshold, keeping the choice arbitrary among equivalents:

```{r prune}
pruned <- ldPruneByCorrelation(X, corThres = 0.8)
ncol(pruned$X.new)
pruned$filter.id
```

It returns the reduced matrix and the retained column indices, so a parallel
table of statistics can be subset the same way.

`ldClumpByScore()` keeps the *best-scoring* variant in each LD clump rather
than an arbitrary one — pass PIPs or −log10 p-values as `score` and it
behaves like PLINK's clumping:

```{r clump, eval=FALSE}
ldClumpByScore(X, score = pips, chr = chrom, pos = position, r2 = 0.2)
```

`enforceDesignFullRank()` is the targeted version: rather than thresholding
everything, it drops variants iteratively until the design plus its
covariates reaches full column rank, and stops there.

```{r full-rank}
Xdup <- cbind(X, dup = X[, 1]) # exactly collinear
dim(enforceDesignFullRank(Xdup, C = matrix(1, nrow(Xdup), 1)))
```

The `C` argument is the covariate matrix the design will sit alongside —
rank has to be assessed on the combined matrix, since a variant can be
independent of the other variants yet collinear with a covariate.
`strategy` chooses what to sacrifice: the most correlated variant, the
lowest-variance one, or the one least correlated with the response.

## Aligning summary statistics to a panel

The counterpart to all of this is on the summary-statistics side: variants
must exist in the panel, and their alleles must be oriented the same way.
`filterVariantsByLdReference()` and `mergeVariantInfo()` do that, and are
covered in the [QC vignette](rss-qc.html), where the motivation lives.

## Next steps

- [QC for summary statistics](rss-qc.html) — matching statistics to a panel.
- [Heritability](heritability.html) — `buildLdEigen()` / `buildLdScore()`
  turn per-block LD into the statistics the h2 estimators consume.
- [Fine-mapping](fine-mapping.html) and [TWAS weights](twas-weights.html) —
  the RSS methods that consume LD.

# Session info

```{r session-info}
sessionInfo()
```
