---
title: "Multivariate adaptive shrinkage across contexts with mash"
author: "pecotmr authors"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multivariate adaptive shrinkage across contexts with mash}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
hasMash <- requireNamespace("mashr", quietly = TRUE) &&
    requireNamespace("flashier", quietly = TRUE)
knitr::opts_chunk$set(
    echo = TRUE,
    message = FALSE,
    warning = FALSE,
    fig.width = 7,
    fig.height = 4.5,
    eval = hasMash
)
```

```{r mash-missing, echo=FALSE, eval=!hasMash, results="asis"}
cat(
    "**Note:** `mashr` and/or `flashier` are not installed, so the code",
    "below is shown but not run."
)
```

## Overview

A QTL measured in several contexts — tissues, cell types, conditions — is
usually not a set of independent experiments. Effects are shared, and an
effect that looks null in one context is often a real effect that context
was underpowered to see.

[mash](https://github.com/stephenslab/mashr) learns the covariance structure
of effect sharing from the data and shrinks each variant's per-context
estimate toward it. `mashPipeline()` drives the whole fit; the chainable
steps behind it let you swap out any stage.

The input is a named list of summary-statistic collections split into
**partitions**, each with a distinct job:

| Partition | Role |
|---|---|
| `strong` | discovery variants; the data-driven covariances are learned here |
| `random` | unbiased background; the mixture weights are fit here |
| `null` | null variants; the residual correlation is estimated here |

`strong` and `random` are required. `null` is optional and used only by the
residual-correlation estimators that need it.

Fitting chunks below are run with `results = "hide"`: `mashr` and
`flashier` narrate their progress on stdout, which knitr cannot filter, so
each fit is split into a compute chunk and a display chunk. The code is
otherwise exactly what you would type.

```{r load-packages}
library(pecotmr)
data(qtlSumStatsMulticontextExample)
ss <- qtlSumStatsMulticontextExample
ss
```

## The one-shot fit

```{r mash-pipeline, results="hide"}
sumStatsList <- list(strong = ss, random = ss, null = ss)
fit <- mashPipeline(sumStatsList, alpha = 0, nPcs = 2L)
```

```{r mash-pipeline-show}
length(fit$U)
round(head(sort(fit$w, decreasing = TRUE), 5), 4)
```

`fit$U` is the prior covariance list and `fit$w` the estimated mixture
weights over it. A large weight on `null` is normal and healthy: most
variants are not QTLs.

`alpha` must match the scale you are on: `alpha = 0` for effect sizes
(BETA/SE), `alpha = 1` for Z-scores. `inputScale` controls which pair is
extracted from the sumstats and defaults to `"auto"` — BETA + SE when every
entry carries both, otherwise Z with unit standard errors.

## Residual correlation

Conditions measured on overlapping samples have correlated residuals, and
mash needs an estimate of that correlation (`Vhat`) to avoid reading shared
noise as shared signal. `residualCorrelationMethod` chooses the estimator:

| Value | Estimator | Needs |
|---|---|---|
| `"identity"` (default) | no residual correlation | — |
| `"simple"` | `mashr::estimate_null_correlation_simple()` | `null` |
| `"simpleSpecific"` | nearest PD correlation of `cov(nullZ)` | `null` |
| `"corshrink"` | adaptive shrinkage of the null correlation | `null`, CorShrink |
| `"mle"` | EM refinement against the prior | `random`, `priorCovariances` |

The default is `"identity"` because it always works, **not** because it is
the best choice. Multi-context QTL data is the overlapping-sample case par
excellence — the same donors appear in every context — so the identity
assumption is usually wrong. Prefer a data-driven estimate whenever you have
a null set:

```{r mash-vhat, results="hide"}
fitSimple <- mashPipeline(
    sumStatsList,
    alpha = 0,
    nPcs = 2L,
    residualCorrelationMethod = "simple"
)
```

```{r mash-vhat-show}
round(head(sort(fitSimple$w, decreasing = TRUE), 5), 4)
```

Supplying a `null` partition while leaving the default in place reports that
the partition is going unused, since that combination is more often an
oversight than an intent.

You can also skip estimation entirely and pass a `Vhat` computed elsewhere —
useful when it was estimated on a larger reference and shipped as a static
artefact:

```{r mash-vhat-supplied, eval=FALSE}
mashPipeline(sumStatsList, alpha = 0, residualCorrelation = myVhat)
```

## The chain, step by step

`mashPipeline()` is a wrapper over four steps. Take them apart when you want
to reuse an intermediate, try several priors against one `Vhat`, or compute
posteriors for a set other than the one you fit on.

```{r mash-chain, results="hide"}
vhat <- mashResidualCorrelation(sumStatsList, alpha = 0, method = "simple")
prior <- mashPriorCovariances(sumStatsList, alpha = 0, vhat = vhat, nPcs = 2L)
model <- mashModelFit(
    sumStatsList,
    alpha = 0,
    priorCovariances = prior,
    vhat = vhat
)
```

```{r mash-chain-show}
class(model)
```

`mashPriorCovariances()` returns `list(U, w, loglik)`, while
`mashCovarianceComponents()` returns a bare covariance list. Either can be
handed to any consumer — the wrapper is unwrapped for you, so there is no
need to reach for `$U`.

`mashModelFit()` returns a `mashr` model object. That is not the same thing
as the `MashPrior` class, which packages a prior for use as a TWAS weight
method; see [TWAS weights](twas-weights.html).

## Posteriors

`mashPosterior()` applies the fitted model to a target set. Fitting on
`random` and computing posteriors on `strong` is the standard split — the
weights should not be learned from the same variants they are applied to.

```{r mash-posterior, results="hide"}
post <- mashPosterior(model, ss, alpha = 0, vhat = vhat)
```

```{r mash-posterior-show}
names(post)
dim(post$PosteriorMean)
```

| Matrix | Meaning |
|---|---|
| `PosteriorMean` / `PosteriorSD` | shrunk effect and its uncertainty |
| `lfsr` | local false sign rate — the quantity to threshold on |
| `lfdr` | local false discovery rate |
| `NegativeProb` | posterior probability the effect is negative |
| `PosteriorCov` | per-variant across-context covariance |

`lfsr` is the usual significance measure here: it asks whether the *sign* of
an effect is confidently known, which is a more honest question than whether
it is exactly zero.

```{r mash-lfsr}
# Variants significant in at least one context at lfsr < 0.05
sum(apply(post$lfsr, 1, min) < 0.05)
```

## Choosing the covariance components

The data-driven part of the prior is built from `components` and then
refined by an `engine`:

```{r mash-components, eval=FALSE}
mashPriorCovariances(
    sumStatsList,
    alpha = 0,
    vhat = vhat,
    components = c("canonical", "pca", "flash", "flashNonneg"),
    engine = "cov_ed", # or "ud" / "ud_ted" (udr)
    nPcs = 2L
)
```

`"cov_ed"` is the default and is mashr's own extreme deconvolution. The
`udr` engines are opt-in and have known numerical issues; `"ud_ted"`
additionally assumes i.i.d. (Z-scale) data. Their dominant cost is
`udControl$n_unconstrained` (default 50), which is worth cutting sharply
when you have few conditions.

To bypass the covariance chain entirely and fit weights against a prior you
already have:

```{r mash-prior-supplied, eval=FALSE}
mashPipeline(sumStatsList, alpha = 0, priorCovariances = myU)
```

## Posterior contrasts and feature scores

Posterior means answer "how big is the effect in each context". A different
question — "which contexts does this feature deviate in, relative to its own
average" — is what the contrast layer computes.

```{r mash-contrast, results="hide"}
data(mashPosteriorExample)
p <- mashPosteriorExample
contrast <- mashPosteriorContrast(
    p$posterior$PosteriorMean,
    p$posterior$PosteriorCov,
    p$orig$bhat
)
scores <- calculateFeatureScores(contrast)
```

```{r mash-contrast-show}
scores
```

Each row is one condition's meta-analysed deviation score across features —
a per-context summary of how distinctive that context's effects are. The
bundled example is a brain cell-type panel, so the conditions are cell
types (`Ast`, `Exc`, `Mic`, ...) plus the `ALL` pooled column.

## Three ways to score a feature

`calculateFeatureScores()` is one of three scoring routes over the same
contrast result, and they answer different questions:

```{r scores-three}
# 1. meta-analysed deviation z, across features (shown above)
head(scores, 3)

# 2. the fraction of features whose deviation is significant
nSignificantScore(contrast, pCutoff = 1e-5)
```

`nSignificantScore()` asks *how often* a condition deviates rather than *how
strongly* — a condition can carry a large mean deviation driven by a handful
of features, or a modest one that is consistent across many, and only the
second is captured here.

The third route brings fine-mapping in. `scoreFromCs()` restricts to the
lead variant of each credible set, so the score reflects deviation at
variants that are plausibly causal rather than at whatever tagged them:

```{r score-cs}
scoreFromCs(p$fineMapping, contrast, condition = "Exc")
```

It takes the least-significant credible-set lead and reports the largest
pairwise |mean|/se there — a deliberately conservative summary, since one
strong credible set should not carry a condition on its own.

The contrast design itself is built from `makePairwiseContrastCol()`, which
sets +1 and −1 for a pair of conditions in a zero vector, and
`fitMashContrast()`, which fits one feature's contrast. `mashPosteriorContrast()`
calls both across all features and pairs, so you only need them directly when
building a non-pairwise design.

## Utilities

The mash track ships a set of helpers for getting real data into shape:

| Function | Use |
|---|---|
| `mashInput()` | build the `(strong, random, null)` matrices from fine-mapping output |
| `sanitizeMashData()` | replace `NaN`/`Inf` in `bhat`/`sbhat` with safe values |
| `filterInvalidSummaryStat()` | drop rows mash cannot use |
| `mergeMashData()` / `sliceMashData()` | combine or subset the matrix bundles |
| `mashRandNullSample()` | draw the random/null partitions |
| `filterMixtureComponents()` | drop components with negligible weight |
| `metaAnalysisPerCondition()` | per-condition fixed-effect meta-analysis |
| `updateMashModelCov()` | swap a fitted model's covariance list |

## Next steps

- [Building QtlSumStats objects](constructing-sumstats.html) — assembling
  the multi-context input.
- [TWAS weights](twas-weights.html) — where a mash prior is consumed, via
  `MashPrior` and mr.mash.

# Session info

```{r session-info}
sessionInfo()
```
