---
title: "Imputing Missing Values in Molecular Trait Matrices"
author: "pecotmr authors"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Imputing Missing Values in Molecular Trait Matrices}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5)
```

## Overview

Real molecular trait data (expression, splicing, protein abundance,
chromatin accessibility) almost always has missing values: low-coverage
samples, transcripts below detection in some tissues, batch dropouts.
Many downstream pecotmr workflows expect a complete matrix:

- [Fine-mapping](fine-mapping.html) — `univariate_analysis_pipeline()`
  requires a numeric `Y` vector with no NAs.
- [TWAS weights](twas-weights.html) — `twas_weights_pipeline()` and
  `twas_multivariate_weights_pipeline()` require complete `Y`.
- Multivariate fits — `mrmash_wrapper()` and `multivariate_analysis_pipeline()`
  reject NA in `Y`.

`pecotmr` provides three imputation methods plus a unified pipeline.
The methods differ in what they assume about the data:

| Method | Function | Assumption | When to use |
|---|---|---|---|
| **XGBoost** | `xgboost_imputation()` | None — fits arbitrary nonlinear per-column models | Heterogeneous matrices where columns have different relationships. Slowest but most flexible. |
| **softImpute** | `softimpute_imputation()` | Low-rank smooth signal | Expression matrices with a small number of dominant axes of variation. Fastest. |
| **flashier** | `flashier_imputation()` | Sparse + scale-mixture EBMF factors | Matrices with sparse + diffuse signal mixtures. Good middle ground; reuses the same prior families as `compute_cov_flash()` for mr.mash priors. |

All three are exposed through `molecular_trait_imputation_pipeline()`,
which can fit multiple methods in one call and evaluate them on a
held-out random mask of observed entries.

```{r load-packages}
library(pecotmr)
```

## Quick start: one method on a synthetic matrix

A low-rank synthetic example illustrates the contract. We build a
60 × 15 matrix with rank-3 truth + Gaussian noise, then mask 10% of
entries.

```{r make-synthetic}
set.seed(1)
n <- 60; p <- 15; rank_true <- 3
U <- matrix(rnorm(n * rank_true), n, rank_true)
V <- matrix(rnorm(p * rank_true), p, rank_true)
truth <- U %*% t(V) + matrix(rnorm(n * p, sd = 0.1), n, p)
rownames(truth) <- paste0("S", seq_len(n))
colnames(truth) <- paste0("F", seq_len(p))

# Mask 10% of entries
na_mask <- matrix(runif(n * p) < 0.1, n, p)
X <- truth
X[na_mask] <- NA
sprintf("Missingness: %.1f%%", 100 * mean(is.na(X)))
```

Run the three methods individually:

```{r single-method-runs}
Xi_softimpute <- softimpute_imputation(X)
Xi_flashier   <- flashier_imputation(X, verbose = 0)
# XGBoost can be slow; reduce iterations for the vignette.
Xi_xgboost    <- xgboost_imputation(X, maxiter = 3, nrounds = 20,
                                     verbose = FALSE)

# RMSE against the truth on the masked cells
rmse <- function(Xi) sqrt(mean((Xi[na_mask] - truth[na_mask])^2))
cat(sprintf("softImpute RMSE: %.3f\n", rmse(Xi_softimpute)))
cat(sprintf("flashier   RMSE: %.3f\n", rmse(Xi_flashier)))
cat(sprintf("xgboost    RMSE: %.3f\n", rmse(Xi_xgboost)))
```

On rank-3 truth, flashier and softImpute both exploit the low-rank
structure; XGBoost has no structural assumption and pays for that on
this kind of input. Real expression matrices tend to be closer to the
low-rank regime, so softImpute and flashier are usually competitive.

## The pipeline: multi-method + held-out CV

For real data you typically want to compare methods before picking one.
`molecular_trait_imputation_pipeline()` does this in a single call.
With `evaluate = TRUE` it holds out a random fraction of observed cells,
fits each method against the masked matrix, and reports RMSE on the
masked cells.

```{r pipeline-multi}
res <- molecular_trait_imputation_pipeline(
  X,
  methods       = c("xgboost", "softimpute", "flashier"),
  method_args   = list(
    xgboost = list(maxiter = 3, nrounds = 20, verbose = FALSE)
  ),
  evaluate      = TRUE,
  eval_fraction = 0.05,
  eval_seed     = 7
)

res$missingness$fraction          # overall missingness
res$evaluation                     # per-method RMSE + time
str(res$imputed, max.level = 1)    # final imputed matrices, one per method
```

The pipeline reports three things:

- **`imputed`** — named list of imputed matrices, one per requested
  method. These were fit on the **full** input (including the cells
  that were held out for CV), so they are what you would use
  downstream.
- **`missingness`** — per-column NA counts and overall NA fraction.
- **`evaluation`** — when `evaluate = TRUE`, a data.frame with one row
  per method: `method`, `rmse`, `n_held_out`, `time_seconds`. Use it to
  pick the method that best handles your data's structure.

## Choosing a method

| Method | Typical RMSE behaviour | Cost | Notes |
|---|---|---|---|
| **softimpute** | Best on smooth low-rank matrices. `lambda` is auto-tuned by held-out CV by default. | Very fast (seconds) plus the lambda-grid pass. | Pass a scalar `lambda` to skip tuning; pass a vector to restrict the CV grid. |
| **flashier** | Best on mixed sparse + diffuse signal. Default priors (`ebnm_normal` + `ebnm_normal_scale_mixture`) work well for most expression-like matrices. | Moderate (seconds-to-minutes). | Reuses the same FLASH defaults as `compute_cov_flash()` (mr.mash prior step). |
| **xgboost** | Robust to non-linear column relationships. Can overfit on small matrices. | Slowest (minutes-to-hours for large matrices). | Set `num_workers > 1` for parallel per-column fits. Tune `max_depth` and `nrounds`. |

A simple rule of thumb: run all three with `evaluate = TRUE` on a
small subset of your data, then use the best for the full matrix.

## Integration with the rest of pecotmr

The imputed matrix slots directly into the downstream pipelines:

```{r downstream, eval=FALSE}
# Imputed phenotype matrix from the pipeline
Y_complete <- res$imputed[["flashier"]]

# Fit fine-mapping on a single context (one column of Y)
fit <- univariate_analysis_pipeline(
  X = genotype_matrix, Y = Y_complete[, "F1"], maf = maf_vector,
  twas_weights = TRUE
)

# Or fit multivariate models on the full matrix
mnm_fit <- multivariate_analysis_pipeline(
  X = genotype_matrix, Y = Y_complete,
  data_driven_prior_matrices = prior_matrices
)
mt <- twas_multivariate_weights_pipeline(
  X = genotype_matrix, Y = Y_complete, mnm_fit = mnm_fit
)
```

Imputation happens once, upstream of fine-mapping and TWAS weight
training, on the full phenotype matrix across contexts. The pipelines
do not re-impute internally.

## Caveats

1. **Don't over-impute.** A column that's missing in most samples won't
   recover useful structure no matter which method you use. Drop
   columns with high missingness (`> 50%` is a common cutoff) before
   imputing.
2. **All-NA columns are dropped.** Each method removes columns where
   every entry is missing and emits a one-line message. The returned
   matrix has fewer columns than the input in that case.
3. **CV evaluation uses one random mask.** For more stable RMSE
   estimates, run the pipeline a few times with different `eval_seed`
   values and average.
4. **softImpute `lambda` is auto-tuned by default.** A random
   fraction of observed cells (default 5%) is held out, softImpute is
   fit across a geometric grid of lambdas via warm-start, and the
   lambda with the lowest held-out RMSE wins. Pass `lambda = <scalar>`
   to skip tuning; pass `lambda = <vector>` to restrict the grid.
   The chosen lambda and the full grid scores are attached to the
   returned matrix as the `"softimpute_lambda"` and
   `"softimpute_lambda_cv"` attributes.

```{r session-info}
sessionInfo()
```
