---
title: "Quality Control for GWAS and QTL Summary Statistics"
author: "pecotmr authors"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quality Control for GWAS and QTL Summary Statistics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5)
```

## Overview

Regression-with-summary-statistics (RSS) methods — both fine-mapping and regularized regression — all require matching variants in a summary-statistics table to an LD reference panel. Two are frequently encountered in practice:

1. **Allele alignment.** The sumary statisticss and LD panel may flip the reference and alternate allele, encode
   strand-ambiguous variants inconsistently, or contain indels and
   duplicates that need filtering.
2. **LD-summary-statistic mismatch.** Even when alleles align, individual
   z-scores can be inconsistent with the LD pattern around them because
   of imputation errors, ancestry differences between the GWAS and LD
   panel, or per-variant sample-size differences.

The `summaryStatsQc()` function provides a convenient interface to performing many QC tasks to address these issues. It takes a
`GwasSumStats` or `QtlSumStats` object (see the *Building QtlSumStats and GwasSumStats objects* vignette), and returns a new collection of the same class with updated entries and
a populated `qcInfo` slot. Downstream pipelines require the `qcInfo` to be populated, so QC is mandatory before
fine-mapping, TWAS-weight learning, colocalization, etc.

```{r load-packages}
library(pecotmr)
```

## Running QC

`summaryStatsQc()` operates on a `GwasSumStats` or `QtlSumStats` object
See *Constructing QtlSumStats and GwasSumStats objects* vignette for details on how to make these objects.
In this example the object is `GwasSumStats` object called `gws`, but the interface is the same for `QtlSumStats`.

`summaryStatsQc()` provides many options for the user to tweak:

```{r run-qc, eval=FALSE}
gws_qcd <- summaryStatsQc(
  gws,
  useDbsnpRefCheck      = FALSE,   # opt-in to MungeSumstats dbSNP checks
  removeIndels          = FALSE,
  removeStrandAmbiguous = TRUE,
  mafCutoff             = 0,       # 0 = off; > 0 requires MAF column
  infoCutoff            = 0,       # 0 = off; > 0 requires INFO column
  nCutoff               = 5,
  skipRegion            = NULL,    # e.g. MHC "chr6:25000000-35000000"
  zMismatchQc           = "slalom", # "none", "slalom", "dentist"
  alleleFlipKriging     = FALSE,    # opt-in pre-SLALOM LD-consistency check
  impute                = FALSE,    # RAISS imputation against ldSketch
  matchMinProp          = 0)
```

`gws_qc` is a new `GwasSumStats` of the same class, with cleaned
entries and a populated `qcInfo` slot. The per-entry audit record is
accessible via `getQcInfo()`:

```{r inspect-qcinfo, eval=FALSE}
qc <- getQcInfo(gws_qcd)
str(qc$entryAudit[[1]])    # per-entry counts: variantsIn, mungeSumstatsDropped,
                           # keepVariantsDropped, skipRegionDropped, matchedAgainstSketch,
                           # krigingOutliersDropped, ldMismatchOutliersDropped, ...
qc$options                 # the curated knobs you passed (frozen for provenance)
```

A non-zero `mafCutoff` requires every entry to carry a `MAF` mcol;
non-zero `infoCutoff` requires `INFO`; non-zero `nCutoff` requires `N`.
A missing column with a non-zero cutoff will produce an error - if you don't want a filter to run, leave its
cutoff at the default value of 0.

## Allele-flip QC: the SuSiE kriging prefilter

`alleleFlipKriging = TRUE` enables an optional prefilter that runs **before**
the SLALOM / DENTIST z-mismatch step. Where SLALOM and DENTIST look for
z-scores that are broadly inconsistent with the LD pattern, the kriging
prefilter specifically targets **allele-flip / LD-mismatch outliers** —
variants whose observed z-score disagrees with the value the LD structure
*predicts* for them from their neighbours.

It uses `susieR`'s kriging RSS diagnostic:

1. `susieR::estimate_s_rss()` estimates the overall LD-mismatch scale `s`
   between the harmonized z-scores and the reference LD matrix `R`.
2. `susieR::kriging_rss(z, R, n, s)` computes, for each variant, the
   leave-one-out **conditional distribution** of `z_i` given all the other
   z-scores under the LD model — i.e. the z-score the LD panel predicts for
   that variant.
3. The standardized residual between observed and predicted (`z_std_diff`) is
   ~`N(0, 1)` when the z-scores and LD panel agree. A large residual flags an
   allele-flip / LD-mismatch outlier; variants below a two-sided p-value
   threshold (default `5e-8`) are dropped. The number removed is recorded in
   the `qcInfo` audit as `krigingOutliersDropped`.

Because it needs a per-region LD matrix and the kriging diagnostic, this is an
**RSS-only** check: it draws `R` from the collection's `ldSketch` and requires
a `susieR` that exports `kriging_rss()` / `estimate_s_rss()` (otherwise it
errors — which is why it is off by default).

The same diagnostic is exposed standalone as `krigingOutlierQc()` when you want
to inspect the per-variant predicted z, residual, statistic, and p-value
directly:

```{r kriging, eval=FALSE}
kr <- krigingOutlierQc(zScore = z, R = ld, n = 10000)
kr$outlier              # logical vector: TRUE = flagged as an outlier
head(kr$diagnostics)    # variant_id, z, predicted, residual, statistic,
                        # p_value, outlier
```

Kriging and SLALOM/DENTIST are complementary and can be combined: the kriging
prefilter removes allele-flip outliers up front, then `zMismatchQc` handles the
remaining LD-mismatch variants.

## LD mismatch correction with SLALOM and DENTIST

`summaryStatsQc(zMismatchQc = ...)` dispatches to one of two
LD-mismatch detectors. They differ in how they decide whether a
variant's z-score is inconsistent with the LD pattern around it.

**SLALOM** (Kanai et al. 2022, the recommended default once you turn
zMismatchQc on):

- For each variant, compares its z-score against the *lead* variant's
  z-score scaled by their LD: a large `(z_i − r_{i,lead} · z_lead)`
  flags the variant as inconsistent.
- Restricted to variants in high LD with the lead (`r2Threshold`).
- Stable: results do not depend on window-size parameters.

**DENTIST** (Chen et al. 2021):

- Iteratively imputes each z-score from all other z-scores via SVD-
  truncated LD regression, then flags large
  `(z_obs − z_imp)^2 / (1 − r²)` values as outliers.
- Operates on windows of the chromosome; the window size noticeably
  affects how many outliers it reports.

Both methods are available directly as `slalom()` / `dentist()` /
`dentistSingleWindow()` if you want to run them outside the
`summaryStatsQc()` pipeline (typically to tune parameters or to inspect
the per-variant intermediate output).

### Dentist combines imputation and LD mismatch correction

DENTIST's outlier rate is sensitive to its window size. On a typical
∼17k-variant test region we observe:

| `min_dim` | Approx outlier rate |
|---|---|
| 2,000 | ∼8 % |
| 6,000 | ∼1.5 % |
| 20,000 | ∼0.6 % |

For this reason SLALOM is the recommended default when you turn
zMismatchQc on. Use DENTIST when you specifically want its iterative
imputation output, or when you are reproducing a DENTIST-based
pipeline.

### Combining LD mismatch correction and imputation with SLALOM and RAISS

SLALOM only *flags* outliers; it does not impute them. To re-impute the
flagged variants from the surviving ones, pair SLALOM with **RAISS** (a
LD-regression imputer). `summaryStatsQc(zMismatchQc = "slalom", impute = TRUE)`
does this for you. The standalone version is exposed as `raiss()`; the
full set of RAISS knobs lives under the `imputeOpts` argument to
`summaryStatsQc()`:

```{r raiss-args, eval=FALSE}
gws_qcd <- summaryStatsQc(
  gws,
  zMismatchQc = "slalom",
  impute      = TRUE,
  imputeOpts  = list(rcond = 0.01, r2Threshold = 0.6,
                     minimumLd = 5, lamb = 0.01))
```


## Next steps

`QtlSumStats` and `GwasSumStats` are direct inputs to
`fineMappingPipeline()`, `twasWeightsPipeline()` (`QtlSumStats` only) and `colocBoostPipeline()`— see the *Fine-mapping with pecotmr*, *Learning TWAS weights with pecotmr* and *Multi-trait colocalization with ColocBoost* vignettes.

```{r session-info}
sessionInfo()
```
