---
title: "Quality Control for GWAS and QTL Summary Statistics"
author: "pecotmr authors"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quality Control for GWAS and QTL Summary Statistics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    message = FALSE,
    warning = FALSE,
    fig.width = 7,
    fig.height = 4.5
)
```

## Overview

Regression-with-summary-statistics (RSS) methods — both fine-mapping and
regularized regression — all require matching variants in a summary-statistics
table to an LD reference panel. Two are frequently encountered in practice:

1.  **Allele alignment.** The sumary statisticss and LD panel may flip the
    reference and alternate allele, encode strand-ambiguous variants
    inconsistently, or contain indels and duplicates that need filtering.
2.  **LD-summary-statistic mismatch.** Even when alleles align, individual
    z-scores can be inconsistent with the LD pattern around them because of
    imputation errors, ancestry differences between the GWAS and LD panel, or
    per-variant sample-size differences.

The `summaryStatsQc()` function provides a convenient interface to performing
many QC tasks to address these issues. It takes a `GwasSumStats` or
`QtlSumStats` object (see the *Building QtlSumStats and GwasSumStats objects*
vignette), and returns a new collection of the same class with updated entries
and a populated `qcInfo` slot. Downstream pipelines require the `qcInfo` to be
populated, so QC is mandatory before fine-mapping, TWAS-weight learning,
colocalization, etc.

```{r load-packages}
library(pecotmr)
```

## Running QC

`summaryStatsQc()` operates on a `GwasSumStats` or `QtlSumStats` object See the
*Building QtlSumStats and GwasSumStats objects* vignette for details on how to
make these objects. In this example the object is `GwasSumStats` object called
`gws`, but the interface is the same for `QtlSumStats`.

`summaryStatsQc()` provides many options for the user to tweak:

```{r run-qc, eval=FALSE}
gws_qcd <- summaryStatsQc(
    gws,
    useDbsnpRefCheck = FALSE, # opt-in to MungeSumstats dbSNP checks
    removeIndels = FALSE,
    removeStrandAmbiguous = TRUE,
    mafCutoff = 0, # 0 = off; > 0 requires MAF column
    infoCutoff = 0, # 0 = off; > 0 requires INFO column
    nCutoff = 5,
    skipRegion = NULL, # e.g. MHC "chr6:25000000-35000000"
    zMismatchQc = "slalom", # "none", "slalom", "dentist"
    alleleFlipKriging = FALSE, # opt-in pre-SLALOM LD-consistency check
    impute = FALSE, # RAISS imputation against ldSketch
    matchMinProp = 0
)
```

`gws_qc` is a new `GwasSumStats` of the same class, with cleaned
entries and a populated `qcInfo` slot. The per-entry audit record is
accessible via `getQcInfo()`:

```{r inspect-qcinfo, eval=FALSE}
qc <- getQcInfo(gws_qcd)
str(qc$entryAudit[[1]]) # per-entry counts: variantsIn, mungeSumstatsDropped,
# keepVariantsDropped, skipRegionDropped, matchedAgainstSketch,
# krigingOutliersDropped, ldMismatchOutliersDropped, ...
qc$options # the curated knobs you passed (frozen for provenance)
```

A non-zero `mafCutoff` requires every entry to carry a `MAF` mcol; non-zero
`infoCutoff` requires `INFO`; non-zero `nCutoff` requires `N`. A missing column
with a non-zero cutoff will produce an error - if you don't want a filter to
run, leave its cutoff at the default value of 0.

## The pieces QC is built from

`summaryStatsQc()` orchestrates the steps below. Each is exported, because
the same operations come up outside a full QC run — preparing a variant list,
auditing why something was dropped, or applying one check without the rest.

### Classifying variants

Allele strings decide how a variant is handled: indels cannot be matched to a
panel the same way SNPs are, and strand-ambiguous SNPs need care.

```{r classify}
ids <- c(
    "chr22:10685239:C:T",
    "chr22:10761227:G:A",
    "chr22:11285239:CT:C",
    "chr22:11500000:A:T"
)
classifyVariantType(ids)
```

### Matching a variant list to an LD panel

`filterVariantsByLdReference()` keeps only variants the panel actually
carries, and reports how many it dropped. It returns both the surviving IDs
and their positions in the input, so a parallel table can be subset the same
way.

```{r filter-ld}
ldMeta <- system.file(
    "extdata",
    "ld_reference",
    "ld_meta_file.tsv",
    package = "pecotmr"
)
kept <- filterVariantsByLdReference(ids, ldMeta)
kept$data
kept$idx
```

`keepIndel = FALSE` additionally drops indels, which is what an
RSS method wanting a SNP-only design needs:

```{r filter-ld-snps}
filterVariantsByLdReference(ids, ldMeta, keepIndel = FALSE)$data
```

### Reconciling two variant tables

`mergeVariantInfo()` matches two variant sets while allowing for flipped
alleles — the second table is re-oriented to the first rather than treated as
a different variant:

```{r merge-variants}
panel <- data.frame(
    chrom = "chr22", pos = c(10685239, 10761227),
    ref = c("C", "G"), alt = c("T", "A")
)
sumstats <- data.frame(
    chrom = "chr22", pos = c(10685239, 10761227),
    ref = c("T", "G"), alt = c("C", "A") # first variant is flipped
)
mergeVariantInfo(panel, sumstats)
```

The flipped variant comes back on the panel's orientation. `all = FALSE`
returns only the re-oriented second set instead of the union.

### Applying one mismatch check

`ldMismatchQc()` is the LD-mismatch step on its own, taking z-scores plus
either a correlation matrix (`R`) or genotypes (`X`):

```{r ld-mismatch, eval=FALSE}
ldMismatchQc(zScore = z, R = ldMatrix, nSample = 10000, method = "slalom")
```

and `autoDecision()` is the helper that reads credible-set structure to
decide whether a flagged variant should be imputed or dropped — the choice
`summaryStatsQc()` otherwise makes for you.

## Allele-flip QC: the SuSiE kriging prefilter

`alleleFlipKriging = TRUE` enables an optional prefilter that runs **before**
the SLALOM / DENTIST z-mismatch step. Where SLALOM and DENTIST look for
z-scores that are broadly inconsistent with the LD pattern, the kriging
prefilter specifically targets **allele-flip / LD-mismatch outliers** —
variants whose observed z-score disagrees with the value the LD structure
*predicts* for them from their neighbours.

It uses `susieR`'s kriging RSS diagnostic:

1.  `susieR::estimate_s_rss()` estimates the overall LD-mismatch scale `s`
    between the harmonized z-scores and the reference LD matrix `R`.
2.  `susieR::kriging_rss(z, R, n, s)` computes, for each variant, the
    leave-one-out **conditional distribution** of `z_i` given all the other
    z-scores under the LD model — i.e. the z-score the LD panel predicts for
    that variant.
3.  The standardized residual between observed and predicted (`z_std_diff`) is
    ~`N(0, 1)` when the z-scores and LD panel agree. A large residual flags an
    allele-flip / LD-mismatch outlier; variants below a two-sided p-value
    threshold (default `5e-8`) are dropped. The number removed is recorded in
    the `qcInfo` audit as `krigingOutliersDropped`.

Because it needs a per-region LD matrix and the kriging diagnostic, this is an
**RSS-only** check: it draws `R` from the collection's `ldSketch` and requires
a `susieR` that exports `kriging_rss()` / `estimate_s_rss()` (otherwise it
errors — which is why it is off by default).

The same diagnostic is exposed standalone as `krigingOutlierQc()` when you want
to inspect the per-variant predicted z, residual, statistic, and p-value
directly:

```{r kriging, eval=FALSE}
kr <- krigingOutlierQc(zScore = z, R = ld, n = 10000)
kr$outlier # logical vector: TRUE = flagged as an outlier
head(kr$diagnostics) # variant_id, z, predicted, residual, statistic,
# p_value, outlier
```

Kriging and SLALOM/DENTIST are complementary and can be combined: the kriging
prefilter removes allele-flip outliers up front, then `zMismatchQc` handles the
remaining LD-mismatch variants.

## LD mismatch correction with SLALOM and DENTIST

`summaryStatsQc(zMismatchQc = ...)` dispatches to one of two
LD-mismatch detectors. They differ in how they decide whether a
variant's z-score is inconsistent with the LD pattern around it.

**SLALOM** (Kanai et al. 2022, the recommended default once you turn
zMismatchQc on):

-  For each variant, compares its z-score against the *lead* variant's z-score
    scaled by their LD: a large `(z_i − r_{i,lead} · z_lead)` flags the variant
    as inconsistent.
- Restricted to variants in high LD with the lead (`r2Threshold`).
- Stable: results do not depend on window-size parameters.

**DENTIST** (Chen et al. 2021):

-  Iteratively imputes each z-score from all other z-scores via SVD- truncated
    LD regression, then flags large `(z_obs − z_imp)^2 / (1 − r²)` values as
    outliers.
-  Operates on windows of the chromosome; the window size noticeably affects
    how many outliers it reports.

Both methods are available directly as `slalom()` / `dentist()` /
`dentistSingleWindow()` if you want to run them outside the
`summaryStatsQc()` pipeline (typically to tune parameters or to inspect
the per-variant intermediate output).

### Dentist combines imputation and LD mismatch correction

DENTIST's outlier rate is sensitive to its window size. On a typical
∼17k-variant test region we observe:

| `min_dim` | Approx outlier rate |
|---|---|
| 2,000 | ∼8 % |
| 6,000 | ∼1.5 % |
| 20,000 | ∼0.6 % |

For this reason SLALOM is the recommended default when you turn
zMismatchQc on. Use DENTIST when you specifically want its iterative
imputation output, or when you are reproducing a DENTIST-based
pipeline.

### Combining LD mismatch correction and imputation with SLALOM and RAISS

SLALOM only *flags* outliers; it does not impute them. To re-impute the
flagged variants from the surviving ones, pair SLALOM with **RAISS** (a
LD-regression imputer). `summaryStatsQc(zMismatchQc = "slalom", impute = TRUE)`
does this for you. The standalone version is exposed as `raiss()`; the
full set of RAISS knobs lives under the `imputeOpts` argument to
`summaryStatsQc()`:

```{r raiss-args, eval=FALSE}
gws_qcd <- summaryStatsQc(
    gws,
    zMismatchQc = "slalom",
    impute = TRUE,
    imputeOpts = list(
        rcond = 0.01,
        r2Threshold = 0.6,
        minimumLd = 5,
        lamb = 0.01
    )
)
```


## Next steps

`QtlSumStats` and `GwasSumStats` are direct inputs to `fineMappingPipeline()`,
`twasWeightsPipeline()` (`QtlSumStats` only) and `colocboostPipeline()`— see
the *Fine-mapping with pecotmr*, *Learning TWAS weights with pecotmr* and
*Multi-trait colocalization with ColocBoost* vignettes.

```{r session-info}
sessionInfo()
```
