---
title: "Learning TWAS weights with pecotmr"
author: "pecotmr authors"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Learning TWAS weights with pecotmr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    message = FALSE,
    warning = FALSE,
    fig.width = 7,
    fig.height = 4.5
)
```

## Overview

TWAS weights are per-variant coefficients that predict a molecular
phenotype (gene expression, splicing, protein abundance) from genotype.
Once learned, the weights are combined with GWAS summary statistics to
compute per-gene TWAS Z-scores (covered in the
[TWAS Z-score + causal inference vignette](twas-zscore.html)).

`twasWeightsPipeline()` is a single that accepts multiple input types. The
output is always a `TwasWeights` collection (a `GRangesList` subclass keyed
by `(study, context, trait, method)`, one element per tuple).

| Input class | Methods supported |
|---|---|
| `QtlDataset` | Univariate: LASSO, elastic net, mr.ash, MCP, SCAD, L0Learn, Bayesian alphabet, DPR. Multivariate: mr.mash, mvsusie |
| `MultiStudyQtlDataset` | Iterates over embedded studies; same methods as `QtlDataset` |
| `QtlSumStats` | RSS-only: LASSO, mr.ash, MCP, SCAD, L0Learn, DPR, PRS-CS, P+T. Multivariate: mr.mash, mvsusie |

```{r load-packages}
library(pecotmr)
```

## Bundled inputs

```{r bundled}
data(qtlDatasetExample, qtlSumStatsExample, multiStudyQtlDatasetExample)
```

## Individual-level weights

The simplest individual-level call learns a single regression-based weight set
from a `QtlDataset`. Cross-validation is enabled by default (`cvFolds = 5`).
Like `fineMappingPipeline()`, you need to tell it which genotype window to
extract — `cisWindow` (per trait) is the most common choice.

Fine-mapping methods (`susie`, `susieInf`, `susieAsh`, `mvsusie`, `fsusie`) are
not re-fit by `twasWeightsPipeline()` - instead run fine-mapping models with
`fineMappingPipeline()` first and pass the result via `fineMappingResult = …`
(see the next section).

```{r individual-quickstart}
tw <- twasWeightsPipeline(qtlDatasetExample, methods = "lasso", cisWindow = 1e6)
tw
```

Extract the per-variant weights for one tuple:

```{r individual-weights}
entry <- getTwasWeights(
    tw,
    study = "study1",
    context = "brain",
    trait = "ENSG_example",
    method = "lasso"
)
head(getWeights(entry))
```

CV performance:

```{r individual-cv}
getCvResult(entry)
```

The entry's `variantIds` slot lists the LD-aligned variant set the
weights are defined on:

```{r individual-vids}
length(getVariantIds(entry))
```

## Including a fine-mapping susie fit in the TWAS ensemble

The posterior effect estimates of SuSiE fine-mapping models are themselves a
set of per-variant weight estimates. When you pass an existing
`FineMappingResult` to `twasWeightsPipeline()`, the trimmed susie fit for each
matching `(study, context, trait)` tuple is added to the ensemble as an
additional weights row instead of being re-fit.

```{r individual-with-fmr}
fmr <- fineMappingPipeline(
    qtlDatasetExample,
    methods = "susie",
    cisWindow = 1e6
)
tw2 <- twasWeightsPipeline(
    qtlDatasetExample,
    methods = c("susie", "lasso"),
    cisWindow = 1e6,
    fineMappingResult = fmr
)
```

## Multi-study weights

```{r multistudy}
msTw <- twasWeightsPipeline(
    multiStudyQtlDatasetExample,
    methods = "lasso",
    cisWindow = 1e6
)
table(msTw$study)
```

The call for iterating over the studies of a MultiStudyQtlDataset is identical
to the call for a single QtlDataset.

## Multiple weight methods at once

Pass a character vector to fit several methods on the same fold split and
return them as separate `(method)` rows in the collection. Fine-mapping methods
(`susie`, `susieInf`, `susieAsh`, `mvsusie`) can only be included if their
results are present in the object passed to the `fineMappingResult` argument:

```{r multimethod, eval=FALSE}
# TWAS-regression only — no fineMappingResult needed.
twasWeightsPipeline(
    qtlDatasetExample,
    methods = c("lasso", "enet", "ridge", "mrash")
)

# Add a fine-mapping susie row by providing the FineMappingResult.
fmr <- fineMappingPipeline(
    qtlDatasetExample,
    methods = "susie",
    cisWindow = 1e6
)
twasWeightsPipeline(
    qtlDatasetExample,
    methods = c("susie", "lasso", "enet", "ridge", "mrash"),
    cisWindow = 1e6,
    fineMappingResult = fmr
)
```

## RSS / sumstats weights

```{r sumstats}
twSs <- twasWeightsPipeline(qtlSumStatsExample, methods = "lasso")
twSs
```

The RSS path uses the `ldSketch` genotype panel inside the
`QtlSumStats` collection to compute LD as needed. The same 

## Multivariate (multi-context) weights

When you have multiple contexts for the same trait (e.g. multi-tissue
expression), pass `methods = "mrmash"` to do a joint multivariate fit:

```{r multivariate, eval=FALSE}
twasWeightsPipeline(
    qtlDatasetExample,
    methods = "mrmash",
    contexts = c("brain", "liver")
)
```

For an explicit cross-context joint fit driven by a
`jointSpecification`:

```{r joint, eval=FALSE}
twasWeightsPipeline(
    multiStudyQtlDatasetExample,
    methods = "mrmash",
    jointSpecification = list(axis = "context", contexts = c("brain", "liver"))
)
```

## Calling a weight method directly

Each `methods =` value dispatches to an exported function, and those
functions are usable on their own. Reach for one when you want to fit a
single model to a matrix you already have — tuning a penalty, comparing
methods outside the pipeline, or debugging a method that behaves oddly on
your data.

The individual-level methods take `X` (genotypes) and `y` (phenotype) and
return a weight vector:

```{r direct-method}
set.seed(1)
n <- 200
p <- 30
X <- matrix(rnorm(n * p), n, p, dimnames = list(NULL, paste0("v", 1:p)))
y <- as.vector(X %*% c(rep(0.4, 3), rep(0, p - 3)) + rnorm(n))

w <- lassoWeights(X = X, y = y)
c(length = length(w), nonzero = sum(w != 0))
```

The families, and the `methods =` name each answers to:

| Family | `methods =` | Function |
|---|---|---|
| Penalized regression | `lasso`, `enet`, `mcp`, `scad`, `l0learn` | `lassoWeights()`, `enetWeights()`, … |
| Bayesian alphabet | `bayesA` … `bayesR`, `bLasso` | `bayesAWeights()`, `bLassoWeights()`, … |
| Dirichlet process regression | `dprVb`, `dprGibbs`, `dprAdaptiveGibbs` | `dprVbWeights()`, `dprGibbsWeights()`, … |
| External PRS methods | `prsCs` (sumstats input only) | `prsCs()`, `sdpr()`, `lassosumRss()` |
| mr.mash / mr.ash | `mrmash`, `mrash` | `mrmashWeights()`, `mrashWeights()` |
| SuSiE family | `susie`, `susieInf`, `susieAsh`, `mvsusie`, `fsusie` | `susieWeights()`, `mvsusieWeights()`, … |

Every token is camelCase, matching the function names. Snake_case spellings
are not accepted: `bayes_r`, `dpr_vb` and `susie_inf` all error with the list
of known tokens.

Every family has an RSS counterpart taking summary statistics and an LD
matrix instead of `X`/`y` — `lassosumRssWeights()`, `mcpRssWeights()`,
`susieRssWeights()`, and so on. Those are what
`twasWeightsPipeline()` calls when the input is a `QtlSumStats`.

`prsCs()`, `sdpr()` and `lassosumRss()` are exported alongside their
`*Weights()` adapters because they return more than the coefficient vector.
`sdpr()` also gives a heritability estimate (`h2`), `lassosumRss()` the whole
regularization path (`beta` is variants x lambda, plus the `lambda` grid and
per-fit `loss`), and `prsCs()` the posterior shrinkage parameters (`psiEst`,
`sigmaEst`, `phiEst`). The adapters reduce each of those to the weights the
pipeline consumes, so reach for the raw function when you want the
diagnostics rather than just the weights.

The mr.mash methods need prior covariance matrices, built by
`buildMrmashPriorMatrices()` from components like `computeCovDiag()` and
`computeCovFlash()`.

## Cross-validation, ensembling and prediction

The pipeline's `ensemble` method is not a model of its own — it is a weighted
combination of the others, with the weights chosen by cross-validation.
Those steps are exported too:

- `twasWeightsCv()` runs the cross-validation that scores each method.
- `learnTwasWeights()` fits one method across CV folds.
- `combineTwasWeights()` forms the ensemble from per-method weights.
- `estimateSparsity()` reports how sparse a fitted weight vector is.

Once you have weights, `twasPredict()` applies them to a genotype matrix.
It takes a *named list* of weight vectors, with names ending in `_weights`,
and returns one predicted column per entry:

```{r predict}
pred <- twasPredict(X, list(gene1_weights = w))
names(pred)
round(cor(as.vector(pred[[1]]), y), 3)
```

That correlation is the in-sample prediction accuracy — on real data you
would want the cross-validated version from `twasWeightsCv()` instead, since
in-sample accuracy is optimistic by construction.

## Common parameters

```{r tw-params, eval=FALSE}
twasWeightsPipeline(
    qtlDatasetExample,
    methods = "susie",
    contexts = "brain",
    traitId = "ENSG_example",
    region = "chr22:25000000-26000000",
    cisWindow = 5e5,
    cvFolds = 5,
    samplePartition = NULL,
    maxCvVariants = -1,
    cvThreads = 1,
    ensemble = TRUE,
    ensembleR2Threshold = 0.01,
    ensembleSolver = "quadprog",
    estimatePi = TRUE,
    phenotypeCovariatesToResidualize = NULL,
    genotypeCovariatesToResidualize = NULL,
    dataType = NULL,
    naAction = "drop"
)
```

-  `ensemble = TRUE` adds an additional `ensembleWeights` row to the collection
    — a weighted combination of the individual methods' CV predictions filterd
    by `ensembleR2Threshold`.
-  `naAction` is forwarded to `getResidualizedPhenotypes()` (see the
    [fine-mapping vignette](fine-mapping.html#common-parameters) for
    semantics).

## Reading the result

`TwasWeights` is a `GRangesList` subclass keyed by the 4-tuple
`(study, context, trait, method)` — one element per tuple, holding the
variants its weights are aligned to. The tuple and the per-tuple payloads are
`mcols()` columns, reached with `$`:

```{r tw-shape}
tw
colnames(tw)
tw$method
```

`as.data.frame()` gives the long per-variant view across every method at once,
which is usually what you want for inspection:

```{r tw-table}
head(as.data.frame(tw), 3)
```

To pull one method's weights, pass the tuple to the accessor:

```{r tw-one-tuple}
key <- list(
    study = tw$study[[1L]],
    context = tw$context[[1L]],
    trait = tw$trait[[1L]],
    method = tw$method[[1L]]
)
w <- do.call(getWeights, c(list(tw), key))
length(w)
do.call(getStandardized, c(list(tw), key)) # standardized or original scale
do.call(getDataType, c(list(tw), key)) # tag carried through for reporting
```

`getWeights()` returns a named numeric vector for univariate methods and a
matrix for multivariate ones. The remaining payloads are keyed the same way:
`getVariantIds()` for the aligned variant set, `getCvResult()` for the
cross-validation record (`rsq`, `r`, predicted values), and `getFits()` for
the full fit object when the pipeline was run with `retainFit = TRUE`.
`resolveWeights()` returns the weights and their variant IDs together, which
is the form the downstream TWAS and MR steps consume.

## Next steps

-  Combine the learned weights with a GWAS via
    [`causalInferencePipeline`](twas-zscore.html) to compute per-gene TWAS
    Z-scores.

# Session info

```{r session-info}
sessionInfo()
```
