---
title: "Learning TWAS weights with pecotmr"
author: "pecotmr authors"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Learning TWAS weights with pecotmr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5)
```

## Overview

TWAS weights are per-variant coefficients that predict a molecular
phenotype (gene expression, splicing, protein abundance) from genotype.
Once learned, the weights are combined with GWAS summary statistics to
compute per-gene TWAS Z-scores (covered in the
[TWAS Z-score + causal inference vignette](twas-zscore.html)).

`twasWeightsPipeline()` is a single that accepts multiple input types. The output is always a `TwasWeights` collection (DFrame
subclass keyed by `(study, context, trait, method)` with per-tuple
`TwasWeightsEntry` payloads).

| Input class | Methods supported |
|---|---|
| `QtlDataset` | Univariate: LASSO, elastic net, mr.ash, MCP, SCAD, L0Learn, Bayesian alphabet, DPR. Multivariate: mr.mash, mvsusie |
| `MultiStudyQtlDataset` | Iterates over embedded studies; same methods as `QtlDataset` |
| `QtlSumStats` | RSS-only: LASSO, mr.ash, MCP, SCAD, L0Learn, DPR, PRS-CS, P+T. Multivariate: mr.mash, mvsusie |

```{r load-packages}
library(pecotmr)
```

## Bundled inputs

```{r bundled}
data(qtl_dataset_example,
     qtl_sumstats_example,
     multi_study_qtl_dataset_example)
```

## Individual-level weights

The simplest individual-level call learns a single regression-based weight set from a
`QtlDataset`. Cross-validation is enabled by default (`cvFolds = 5`). Like
`fineMappingPipeline()`, you need to tell it which genotype window to
extract — `cisWindow` (per trait) is the most common choice.

Fine-mapping methods (`susie`, `susieInf`, `susieAsh`, `mvsusie`,
`fsusie`) are not re-fit by `twasWeightsPipeline()` - instead run fine-mapping models with `fineMappingPipeline()` first and pass the result via
`fineMappingResult = …` (see the next section).

```{r individual-quickstart}
tw <- twasWeightsPipeline(qtl_dataset_example,
                           methods   = "lasso",
                           cisWindow = 1e6)
tw
```

Extract the per-variant weights for one tuple:

```{r individual-weights}
entry <- getTwasWeights(tw, study = "study1", context = "brain",
                         trait = "ENSG_example", method = "lasso")
head(getWeights(entry))
```

CV performance:

```{r individual-cv}
getCvResult(entry)
```

The entry's `variantIds` slot lists the LD-aligned variant set the
weights are defined on:

```{r individual-vids}
length(getVariantIds(entry))
```

## Including a fine-mapping susie fit in the TWAS ensemble

The posterior effect estimates of SuSiE fine-mapping models are themselves a set of per-variant
weight estimates. When you pass an existing `FineMappingResult` to
`twasWeightsPipeline()`, the trimmed susie fit for each matching
`(study, context, trait)` tuple is added to the ensemble as an additional
weights row instead of being re-fit.

```{r individual-with-fmr}
fmr <- fineMappingPipeline(qtl_dataset_example, methods = "susie",
                            cisWindow = 1e6)
tw2 <- twasWeightsPipeline(qtl_dataset_example,
                            methods            = c("susie", "lasso"),
                            cisWindow          = 1e6,
                            fineMappingResult  = fmr)
```

## Multi-study weights

```{r multistudy}
msTw <- twasWeightsPipeline(multi_study_qtl_dataset_example,
                             methods   = "lasso",
                             cisWindow = 1e6)
table(msTw$study)
```

The call for iterating over the studies of a MultiStudyQtlDataset is identical to the call for a single QtlDataset.

## Multiple weight methods at once

Pass a character vector to fit several methods on the same fold split
and return them as separate `(method)` rows in the collection.
Fine-mapping methods (`susie`, `susieInf`, `susieAsh`, `mvsusie`) can
only be included if their results are present in the object passed to the `fineMappingResult` argument:

```{r multimethod, eval=FALSE}
# TWAS-regression only — no fineMappingResult needed.
twasWeightsPipeline(
  qtl_dataset_example,
  methods = c("lasso", "enet", "ridge", "mrash"))

# Add a fine-mapping susie row by providing the FineMappingResult.
fmr <- fineMappingPipeline(qtl_dataset_example, methods = "susie",
                            cisWindow = 1e6)
twasWeightsPipeline(
  qtl_dataset_example,
  methods           = c("susie", "lasso", "enet", "ridge", "mrash"),
  cisWindow         = 1e6,
  fineMappingResult = fmr)
```

## RSS / sumstats weights

```{r sumstats}
twSs <- twasWeightsPipeline(qtl_sumstats_example, methods = "lasso")
twSs
```

The RSS path uses the `ldSketch` `GenotypeHandle` inside the
`QtlSumStats` collection to compute LD as needed. The same 

## Multivariate (multi-context) weights

When you have multiple contexts for the same trait (e.g. multi-tissue
expression), pass `methods = "mrmash"` to do a joint multivariate fit:

```{r multivariate, eval=FALSE}
twasWeightsPipeline(
  qtl_dataset_example,
  methods  = "mrmash",
  contexts = c("brain", "liver"))
```

For an explicit cross-context joint fit driven by a
`jointSpecification`:

```{r joint, eval=FALSE}
twasWeightsPipeline(
  multi_study_qtl_dataset_example,
  methods            = "mrmash",
  jointSpecification = list(axis = "context",
                             contexts = c("brain", "liver")))
```

## Common parameters

```{r tw-params, eval=FALSE}
twasWeightsPipeline(
  qtl_dataset_example,
  methods             = "susie",
  contexts            = "brain",
  traitId             = "ENSG_example",
  region              = "chr22:25000000-26000000",
  cisWindow           = 5e5,
  cvFolds             = 5,
  samplePartition     = NULL,
  maxCvVariants       = -1,
  cvThreads           = 1,
  ensemble            = TRUE,
  ensembleR2Threshold = 0.01,
  ensembleSolver      = "quadprog",
  estimatePi          = TRUE,
  phenotypeCovariatesToResidualize = NULL,
  genotypeCovariatesToResidualize  = NULL,
  dataType            = NULL,
  naAction            = "drop")
```

- `ensemble = TRUE` adds an additional `ensembleWeights` row to the
  collection — a weighted combination of the individual
  methods' CV predictions filterd by `ensembleR2Threshold`.
- `naAction` is forwarded to `getResidualizedPhenotypes()` (see the
  [fine-mapping vignette](fine-mapping.html#common-parameters) for
  semantics).

## Reading the result

`TwasWeights` is a DFrame subclass keyed by the 4-tuple
`(study, context, trait, method)`. The per-tuple payload is a
`TwasWeightsEntry`:

```{r tw-entry}
entry <- tw$entry[[1L]]
class(entry)
slotNames(entry)
```

Each entry exposes:

- `weights` — per-variant coefficients (named numeric vector or matrix
  for multivariate methods)
- `variantIds` — variant IDs the weights are aligned to
- `cvResult` — list with `rsq`, `r`, predicted values, etc. (access
  with `getCvResult()`)
- `standardized` — whether weights are on the standardized
  (`y ~ scale(X)`) or original scale
- `dataType` — string tag carried through for reporting (e.g.
  `"expression"`, `"splicing"`)
- `fit` — the full fit object when the user requested `retainFit = TRUE`

## Next steps

- Combine the learned weights with a GWAS via
  [`causalInferencePipeline`](twas-zscore.html) to compute per-gene
  TWAS Z-scores.
