---
title: "Causal inference with pecotmr: TWAS Z-scores and MR"
author: "pecotmr authors"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Causal inference with pecotmr: TWAS Z-scores and MR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    message = FALSE,
    warning = FALSE,
    fig.width = 7,
    fig.height = 4.5
)
```

## Overview

`causalInferencePipeline()` pairs QTL fine-mapping and/or regularized
regression weights one or more GWAS studies to produce, per `(qtlStudy,
context, trait, method, gwasStudy)` tuple:

- a **TWAS Z-score** (and p-value), and
-  when `fineMappingResult` is supplied, a **Mendelian Randomization
    Wald-ratio** estimate over the QTL credible sets.

Three valid input combinations:

| What you supply | TWAS-Z source | MR? |
|---|---|---|
| `twasWeights` only | per-tuple weights from `TwasWeights` | no |
| `fineMappingResult` only | SuSiE coefficients from `getTopLoci()` | yes |
| both | weights from `TwasWeights` | yes (MR uses the FMR) |

```{r load-packages}
library(pecotmr)
```

## Bundled inputs

```{r bundled}
data(qtlDatasetExample, gwasSumStatsS4Example)

# Pre-compute the QTL-side products from the bundled QtlDataset.
fmr <- fineMappingPipeline(
    qtlDatasetExample,
    methods = "susie",
    cisWindow = 1e6
)
tw <- twasWeightsPipeline(
    qtlDatasetExample,
    methods = "susie",
    cisWindow = 1e6,
    fineMappingResult = fmr
)
```

## TWAS Z only (weights alone)

```{r ci-tw-only}
res <- causalInferencePipeline(
    gwasSumStats = gwasSumStatsS4Example,
    twasWeights = tw
)
head(S4Vectors::mcols(res))
```

The result is a `GRanges` object positioned at the variant span of each QTL
weight set, with identity columns (`qtlStudy`, `context`, `trait`,
`method`, `gwasStudy`) and the TWAS columns (`twasZ`, `twasPval`).
MR columns are `NA` in this mode.

## TWAS Z + MR (weights + fine-mapping)

When you also pass `fineMappingResult`, the pipeline adds Wald-ratio MR
on top of the TWAS computation. The MR uses each fine-mapping entry's
`topLoci` rows as candidate instrumental variables.

```{r ci-both}
res2 <- causalInferencePipeline(
    gwasSumStats = gwasSumStatsS4Example,
    twasWeights = tw,
    fineMappingResult = fmr,
    mrMethod = "ivwPerVariant",
    mrPipCutoff = 0.5
)
S4Vectors::mcols(res2)[, c(
    "twasZ",
    "twasPval",
    "waldRatio",
    "waldRatioSe",
    "mrPval",
    "nIV"
)]
```

## MR variants: per-variant vs CS-aware IVW

| `mrMethod` | What it does | Extra filters |
|---|---|---|
| `"ivwPerVariant"` (default) | Treats every `topLoci` variant with `pip > mrPipCutoff` as an IV; pools their Wald ratios via inverse-variance weighting. | `mrPipCutoff` |
| `"csAware"` | Groups variants by credible set (column `cs` in `topLoci`), forms a PIP-weighted composite Wald ratio per CS (subject to per-CS cumulative-PIP cutoff `mrCpipCutoff`), then IVW-pools across CSs. Reports Cochran's Q and I-squared. | `mrCpipCutoff` |

```{r ci-csaware, eval=FALSE}
causalInferencePipeline(
    gwasSumStats = gwasSumStatsS4Example,
    twasWeights = tw,
    fineMappingResult = fmr,
    mrMethod = "csAware",
    mrCpipCutoff = 0.5
)
```

## Cross-method p-value combination

When a `(qtlStudy, context, trait, gwasStudy)` group has TWAS results
from multiple weight methods, `combineMethods` runs
`combinePValues()` to produce a single combined p-value per group.

```{r ci-combine, eval=FALSE}
# Multi-method weights so combination is meaningful
tw_multi <- twasWeightsPipeline(
    qtlDatasetExample,
    methods = c("susie", "lasso")
)
causalInferencePipeline(
    gwasSumStats = gwasSumStatsS4Example,
    twasWeights = tw_multi,
    combineMethods = c("acat", "hmp")
)
```

Accepted values for `combineMethods`: `acat`, `hmp`,
`bonferroni`, `fisher`, `stouffer`, `invchisq`, `gbj`, `bj`, `hc`,
`ghc`, `minp`, `gbj_omni`, `aspu`, `gates` (see `?combinePValues`).

## LD-sketch consistency

Both `twasWeights` and `fineMappingResult` carry an `ldSketch` slot.
When provided, they must reference the same LD panel as
`gwasSumStats`. The pipeline errors on mismatch.

A QTL input with `ldSketch = NULL` (e.g. weights learned from
individual-level data without any RSS step) skips the panel check for
that input — useful for analyses that mix individual-level QTL fits
with sumstats GWAS.

## Common parameters

```{r ci-params, eval=FALSE}
causalInferencePipeline(
    gwasSumStats = gwasSumStatsS4Example,
    twasWeights = tw,
    fineMappingResult = fmr,
    mrPipCutoff = 0.5, # ivwPerVariant
    mrMethod = "ivwPerVariant",
    mrCpipCutoff = 0.5, # csAware
    combineMethods = NULL
) # e.g. c("acat", "hmp")
```

## Reading the result

The `GRanges` output indexes results by the QTL variant span (the
union of variants used in each tuple's weights). Extract per-tuple
values with `S4Vectors::mcols()`:

```{r ci-mcols, eval=FALSE}
mc <- as.data.frame(S4Vectors::mcols(res2))
# Significant TWAS hits at Bonferroni alpha = 0.05 across rows
mc[mc$twasPval < 0.05 / nrow(mc), ]
```

# Session info

```{r session-info}
sessionInfo()
```
