---
title: "Molecular QTL Enrichment Analysis for Complex Traits"
author: "Ru Feng"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Molecular QTL Enrichment Analysis for Complex Traits}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette demonstrates the enrichment analysis part of the `pecotmr`
package, which largely follows from `fastENLOC`
(https://github.com/xqwen/fastenloc) but uses `susieR` fitted objects as input
to estimate priors for use with the `coloc` package (coloc v5, aka
SuSiE-coloc). The main differences are:

1)  enrichment is based on all QTL variants whether or not they are inside
    signal clusters;
2) causal QTL are sampled from SuSiE single effects, not signal clusters;
3)  a variant may be a QTL for not only multiple conditions (e.g. cell types)
    but also multiple regions (e.g. genes).

The example uses de-identified data shipped with the package. All sample names,
variant positions, and identifiers are synthetic.

## Set up environment and load data

```{r load-pkgs}
library(pecotmr)
```

The enrichment pipeline consumes fine-mapping results as `pecotmr` S4 objects:
a `GwasFineMappingResult` for the GWAS and a `QtlFineMappingResult` for the
molecular QTLs. Both are shipped as example datasets.

```{r load-data}
data(gwasFineMappingExample)
data(qtlFineMappingExample)
gwasFineMappingExample
qtlFineMappingExample
```

## Input data

Each object is a `DFrame`-backed collection of per-tuple fine-mapping entries;
use the accessors rather than reaching into slots. `getTopLoci()` returns the
per-variant posterior table (identity columns, `pip`, and credible-set
membership):

```{r toploci}
head(getTopLoci(gwasFineMappingExample))
head(getTopLoci(qtlFineMappingExample))
```

Posterior inclusion probabilities across all variants in each region, taken
from the underlying SuSiE fit:

```{r pip-plots, fig.width = 7, fig.height = 3.2}
gwas_pip <- getSusieFit(gwasFineMappingExample)$pip
qtl_pip <- getSusieFit(qtlFineMappingExample)$pip
op <- par(mfrow = c(1, 2))
plot(
    gwas_pip,
    ylab = "PIP",
    xlab = "Variant index",
    main = "GWAS PIPs",
    pch = 16,
    cex = 0.6
)
plot(
    qtl_pip,
    ylab = "PIP",
    xlab = "Variant index",
    main = "QTL PIPs",
    pch = 16,
    cex = 0.6
)
par(op)
```

## Perform enrichment analysis

`qtlEnrichmentPipeline()` takes the two fine-mapping collections directly and
estimates, for every `(gwasStudy, qtlStudy, qtlContext)` combination, how
enriched the molecular QTLs are among the GWAS causal variants. The GWAS
collection must be RSS-derived (it carries an `ldSketch`);
`gwasFineMappingExample` already does.

```{r enrichment}
enrich_res <- qtlEnrichmentPipeline(
    gwasFineMappingResult = gwasFineMappingExample,
    qtlFineMappingResult = qtlFineMappingExample,
    verbose = FALSE
)
enrich_res
```

The result is a data frame with one row per
`(gwasStudy, qtlStudy, qtlContext)` combination, giving the enrichment estimate
(`enrichment`), its standard error (`enrichmentSe`), and the log-odds
(`enrichmentLogOdds`). These enrichment priors can then be supplied to a
colocalization analysis. This synthetic single-region example is for
illustration only; stable enrichment estimates require many fine-mapped
regions.

## Session information

Here are some details about the computing environment, including the versions
of R and the R packages used to generate these results.

```{r session-info}
sessionInfo()
```
